---
title: "Charactersing effect of anaemia on mortality in severe malaria"
output:
  html_document:
    toc: true
    keep_md: yes
    fig_caption: yes
---

# Preliminaries


```{r, echo=F}
knitr::opts_chunk$set(cache = TRUE, cache.comments = FALSE, 
                      echo = TRUE, include = TRUE, 
                      fig.width = 7, fig.height = 7,
                      fig.pos = 'H', 
                      dev = 'png', dpi = 300)

RUN_MODELS = F
```

```{r, echo=FALSE, include=FALSE}
library(lme4)
require(RColorBrewer)
require(boot)
require(plyr)
require(dplyr)
library(mgcv) # For the GAM modelling
library(epitools) # For risk ratio Wald CIs
library(mice)
library(plotrix)
library(merTools)
```



# Exploratory analysis


This looks at the severe malaria legacy dataset from MORU.

```{r, echo=FALSE}
# original dataset
load('RData/Data.RData')
#imputed datasets
load('RData/Multiple_Imputed_Datasets.RData')

ind_missing = apply(m[,c('AgeInYear','HCT','BUN','BD','LPAR_pct')], 1, function(x) sum(is.na(x))>0)
Leg_data_complete = m[!ind_missing,]
Leg_data_complete = Leg_data_complete[!is.infinite(Leg_data_complete$LPAR_pct),]
Leg_data_complete$studyID = as.factor(Leg_data_complete$studyID)# for plotting
```



First we look at the mortality rates across the different studies.

```{r, echo=FALSE}
for(s in unique(m$studyID)){
  print(paste(s, ', mortality of:', round(100*mean(m$outcome[m$studyID==s])),'%'))
}
```

We look at the quantiles of the ages in the different studies:
```{r, echo=FALSE}
for(s in unique(m$studyID)){
  print(paste0(s, ', ages:', round(quantile(m$AgeInYear[m$studyID==s], 
                                            probs = c(0,.5,1))), 
               collapse = ' '))
}

for(s in unique(m$studyID)){
  print(s)
  print(table(m$drug_class[m$studyID==s]))
}
```


Let's look at the linear associations between the key baseline variables. We use mixed effects linear models to estimate these associations (random effect terms for both country and study).

```{r , echo=FALSE}
mod_Age_HCT = gam(formula = HCT ~ s(AgeInYear)  + s(studyID, bs='re') + 
                    s(country, bs='re'), 
                  data = Leg_data_complete)
mod_HCT_BD = gam(formula = BD ~ s(HCT) + s(studyID, bs='re') + 
                   s(country, bs='re'), 
                 data = Leg_data_complete)
mod_BD_LPAR = gam(formula = BD ~ s(LPAR_pct) + s(studyID, bs='re') + 
                    s(country, bs='re'),
                  data =Leg_data_complete)
mod_BUN_BD = gam(formula = BD ~ s(log10(BUN)) + s(studyID, bs='re') + 
                   s(country, bs='re'),
                 data =Leg_data_complete)
```

```{r ExploratoryPlots, echo=FALSE}
mycols = brewer.pal(length(unique(m$studyID)), 'Set1')
Leg_data_complete$color = as.character(revalue(Leg_data_complete$studyID, 
                                               replace = c('Core Malaria'=mycols[1],
                                                           'AAV'=mycols[2],
                                                           'AQ'=mycols[3],
                                                           'SEAQUAMAT' = mycols[4],
                                                           'AQUAMAT' = mycols[5],
                                                           'QC' = mycols[6])))
par(las=1, mfrow=c(2,2), mar=c(4,4,1,1), bty='n')
## Base Excess and HCT
plot(jitter(Leg_data_complete$HCT,amount = .4), 
     jitter(Leg_data_complete$BD, amount = .4), ylim=c(-6,28),
     col=Leg_data_complete$color, pch='*', xlab='', 
     ylab='Base deficit (mEq/L)', xlim=c(8,47), panel.first = grid())
mtext(text = 'Haematocrit (%)',side = 1,line = 2.5)
ys = predict(object = mod_HCT_BD, 
             newdata = data.frame(HCT=8:50,
                                  studyID='AQ',
                                  country='Vietnam'), 
             exclude = c("s(country)","s(studyID)"))
lines(8:50, ys, lwd=3, col='black')

## Parasitaemia and Anaemia
plot(Leg_data_complete$LPAR_pct, 
     jitter(Leg_data_complete$BD,amount = .4), 
     xlim = c(-2.418006, 1.562435),ylim=c(-6,28),
     col=Leg_data_complete$color, pch='*', 
     xlab='', panel.first = grid(),
     ylab='Base deficit (mEq/L)', xaxt='n')
axis(1, at = log10(c(30,10,1,.1,.01)), labels = c(30,10,1,.1,.01))
mtext(text = 'Parasitised RBCs (%)',side = 1,line = 2.5)
ys = predict(object = mod_BD_LPAR, 
             newdata = data.frame(LPAR_pct=seq(-3,log10(50),by=.1),
                                  studyID='AQ',
                                  country='Vietnam'), 
             exclude = c("s(country)","s(studyID)"))
lines(seq(-3,log10(50),by=.1), ys, lwd=3, col='black')

## BUN and BD
plot(jitter(log10(Leg_data_complete$BUN),amount = .4),
     jitter(Leg_data_complete$BD, amount = .4), ylim=c(-6,28),
     col=Leg_data_complete$color, pch='*', panel.first = grid(),
     xlab='',xlim = c(0.4771213,2.1461280),
     ylab='Base deficit (mEq/L)', xaxt='n')
mtext(text = 'Blood urea nitrogen (mmol/L)',side = 1,line = 2.5)
axis(1, at=c(log10(3), log10(5), 1, log10(30),2), labels = c(3,5,10,30,100))

summary(mod_BUN_BD)
ys = predict(object = mod_BUN_BD, 
             newdata = data.frame(BUN=2:166,
                                  studyID='AQ',
                                  country='Vietnam'), 
             exclude = c("s(country)","s(studyID)"))
lines(log10(2:166), ys, lwd=3, col='black')

## Parasitaemia and Anaemia
plot(jitter(Leg_data_complete$AgeInYear,amount=.4), 
     jitter(Leg_data_complete$HCT,amount = .4), 
     col=Leg_data_complete$color, pch='*', 
     xlab='', ylim=c(8,47), panel.first = grid(),
     ylab='Haematocrit (%)', xlim = c(0,65))
mtext(text = 'Age (years)',side = 1,line = 2.5)
summary(mod_Age_HCT)
ys = predict(object = mod_Age_HCT, 
             newdata = data.frame(AgeInYear=0:80,
                                  studyID='AQ',
                                  country='Vietnam'), 
             exclude = c("s(country)","s(studyID)"))
lines(0:80, ys, lwd=3, col='black')
legend('bottomright',col=mycols,inset = 0.01,
       legend = c('Core Malaria',
                  'AAV',
                  'AQ',
                  'SEAQUAMAT',
                  'AQUAMAT','QC'),pch='*')
```

Effect on survival 

```{r UnAdjusted}
XX = SM_Impute_List[[1]]
kids=XX$AgeInYear<=12
modHCT_kids=gam(outcome ~ s(HCT) + s(studyID, bs='re') + s(country, bs='re'),
           data = XX[kids,], family='binomial')
modHCT_adults=gam(outcome ~ s(HCT) + s(studyID, bs='re') + s(country, bs='re'),
           data = XX[!kids,], family='binomial')

modcoma=gam(outcome ~ coma + s(studyID, bs='re') + s(country, bs='re'),
            data = XX, family='binomial')

modBD=gam(outcome ~ s(BD) + s(studyID, bs='re') + s(country, bs='re'),
          data = XX, family='binomial')

modpoedema=gam(outcome ~ poedema + s(studyID, bs='re') + s(country, bs='re'),
               data = XX, family='binomial')

modconv=gam(outcome ~ convulsions + s(studyID, bs='re') + s(country, bs='re'),
            data = XX, family='binomial')

modshock=gam(outcome ~ shock + s(studyID, bs='re') + s(country, bs='re'),
             data = XX, family='binomial')

modBUN=gam(outcome ~ s(log10(BUN)) + s(studyID, bs='re') + s(country, bs='re'),
           data = XX, family='binomial')

modLPAR_pct=gam(outcome ~ s(LPAR_pct) + s(studyID, bs='re') + s(country, bs='re'),
                data = XX, family='binomial')
```

```{r UnadjustedPlots, echo=FALSE}
par(las=1, bty='n', mfrow=c(2,4))

Factors = c('HCT','coma','BD','poedema','convulsions','BUN')
x_hcts_kids = seq(8,40, by=1)
x_hcts_adults = seq(10,50, by=1)
x_comas = 0:1
x_shock = 0:1
x_conv = 0:1
x_poedema = 0:1
x_bd = seq(-5,35, by=2)
x_BUN = 2:150
x_parasites = seq(-2,log10(30),by=.001)

#************** BD
predsBD = predict(modBD, newdata = data.frame(BD= x_bd, 
                                              studyID='AQ', 
                                              country='Thailand'), 
                  exclude = c("s(country)","s(studyID)"), 
                  se.fit=T)
plot(x_bd, 100*inv.logit(predsBD$fit), ylab='mortality (%)', 
     panel.first = grid(), xlab='', 
     type='l', lwd=3,ylim = c(0,80))
mtext(text = 'Base deficit (mEq/L)',side = 1,line = 2.5, cex=.7)
polygon(c(x_bd, rev(x_bd)), 100*inv.logit(c(predsBD$fit + 1.96*predsBD$se.fit,
                                            rev(predsBD$fit - 1.96*predsBD$se.fit))),
        lty=2,lwd=2, border = NA, col = adjustcolor('blue',alpha.f = .3))

#************** HCT
predsHCTkids = predict(modHCT_kids, 
                   newdata = data.frame(HCT=x_hcts_kids, 
                                        studyID='AQUAMAT', 
                                        country='Kenya'), 
                   exclude = c("s(country)","s(studyID)"), 
                   se.fit=T)
predsHCTadults = predict(modHCT_adults, 
                   newdata = data.frame(HCT=x_hcts_adults, 
                                        studyID='AQ', 
                                        country='Vietnam'), 
                   exclude = c("s(country)","s(studyID)"), 
                   se.fit=T)
plot(x_hcts_kids, 100*inv.logit(predsHCTkids$fit), ylab='mortality (%)', 
     panel.first = grid(), xlab='', xlim = range(c(x_hcts_kids,x_hcts_adults)),
     type='l', lwd=3,ylim = c(0,35),lty='dotted')
lines(x_hcts_adults, 100*inv.logit(predsHCTadults$fit), ylab='mortality (%)', 
     type='l', lwd=3, lty='longdash')
mtext(text = 'Haematocrit (%)',side = 1,line = 2.5,cex=.7)
polygon(c(x_hcts_kids,rev(x_hcts_kids)), 
        100*inv.logit(c(predsHCTkids$fit + 1.96*predsHCTkids$se.fit, 
                        rev(predsHCTkids$fit - 1.96*predsHCTkids$se.fit))),
        lty=2,lwd=2,border = NA, col = adjustcolor('blue',alpha.f = .3))
polygon(c(x_hcts_adults,rev(x_hcts_adults)), 
        100*inv.logit(c(predsHCTadults$fit + 1.96*predsHCTadults$se.fit, 
                        rev(predsHCTadults$fit - 1.96*predsHCTadults$se.fit))),
        lty=2,lwd=2,border = NA, col = adjustcolor('blue',alpha.f = .3))

#************** BUN
predsBUN = predict(modBUN, newdata = data.frame(BUN=x_BUN, 
                                                studyID='AQ', 
                                                country='Thailand'), 
                   exclude = c("s(country)","s(studyID)"), 
                   se.fit=T)
plot(log10(x_BUN), 100*inv.logit(predsBUN$fit), ylab='mortality (%)', 
     xlab='', panel.first = grid(), 
     type='l', lwd=3,ylim = c(0,55), xaxt='n')
mtext(text = 'Blood urea nitrogen (mmol/L)',side = 1,line = 2.5,cex=.7)
axis(1, at = log10(c(2,10,100)), labels = c(2,10,100))
polygon(c(log10(x_BUN),rev(log10(x_BUN))), 
        100*inv.logit(c(predsBUN$fit + 1.96*predsBUN$se.fit,
                        rev(predsBUN$fit - 1.96*predsBUN$se.fit))), 
        lty=2,lwd=2, border = NA, col = adjustcolor('blue',alpha.f = .3))

#************** Log10 parasitaemia
predspara_PCT = predict(modLPAR_pct, newdata = data.frame(LPAR_pct=x_parasites, 
                                                          studyID='AQ', 
                                                          country='Thailand'), 
                        exclude = c("s(country)","s(studyID)"), 
                        se.fit=T)
plot(x_parasites, 100*inv.logit(predspara_PCT$fit), ylab='mortality (%)', 
     xlab='', panel.first = grid(), 
     type='l', lwd=3,ylim = c(0,25), xaxt='n')
mtext(text = 'Parasitized RBCs (%)',side = 1,line = 2.5,cex=.7)
axis(1, at = c(-2,-1,0,log10(30)), labels = c(0.01,0.1,1,30))
axis(1, at = log10(c(10,20)), labels = NA)
polygon(c(x_parasites,rev(x_parasites)), 
        100*inv.logit(c(predspara_PCT$fit + 1.96*predspara_PCT$se.fit,
                        rev(predspara_PCT$fit - 1.96*predspara_PCT$se.fit))), 
        lty=2,lwd=2, border = NA, col = adjustcolor('blue',alpha.f = .3))

#**************  Coma
predscoma = predict(modcoma, newdata = data.frame(coma=x_comas, 
                                                  studyID='AQ', 
                                                  country='Thailand'), 
                    exclude = c("s(country)","s(studyID)"), 
                    se.fit=T)
plot(x_comas, 100*inv.logit(predscoma$fit), ylab='mortality (%)', 
     panel.first = grid(), xlab='', 
     type='p', lwd=3, ylim = c(0,30), xaxt='n', 
     pch=19, xlim = c(-.25,1.25))
axis(1, at=x_comas, labels = c('No Coma','Coma'), 
     tick = F,line = .5,cex.axis=.9)
ys = 100*inv.logit(c(predscoma$fit - 1.96*predscoma$se.fit,
                     predscoma$fit + 1.96*predscoma$se.fit))
arrows(x0 = x_comas[1],y0 = ys[1], x1 = x_comas[1], y1 = ys[3],
       angle = 90,lwd=6,code = 3,length=0,col = adjustcolor('blue',alpha.f = .3))
arrows(x0 = x_comas[2],y0 = ys[2], x1 = x_comas[2], y1 = ys[4],
       angle = 90,lwd=6,code = 3,length=0,col = adjustcolor('blue',alpha.f = .3))


#************** Convulsions
predsconv = predict(modconv, newdata = data.frame(convulsions=x_conv, 
                                                  studyID='AQ', 
                                                  country='Thailand'), 
                    exclude = c("s(country)","s(studyID)"), 
                    se.fit=T)
plot(x_conv, 100*inv.logit(predsconv$fit), 
     panel.first = grid(), ylab='mortality (%)', xlab='', 
     type='p', lwd=3, ylim = c(0,25), xaxt='n',
     pch=19, xlim = c(-.25,1.25))
axis(1, at=x_conv, labels = c('No seizures','Seizures'),
     tick = F,line = .5,cex.axis=.9)

ys = 100*inv.logit(c(predsconv$fit + 1.96*predsconv$se.fit,
                     predsconv$fit -1.96*predsconv$se.fit))
arrows(x0 = x_conv[1],y0 = ys[1], x1 = x_conv[1], y1 = ys[3],angle = 90,
       lwd=6,code = 3,length=0,col = adjustcolor('blue',alpha.f = .3))
arrows(x0 = x_conv[2],y0 = ys[2], x1 = x_conv[2], y1 = ys[4],angle = 90,
       lwd=6,code = 3,length=0,col = adjustcolor('blue',alpha.f = .3))

#************** Pulmonary Oedema
predspoedema = predict(modpoedema, 
                       newdata = data.frame(poedema=x_poedema, 
                                            studyID='AQ', 
                                            country='Thailand'), 
                       exclude = c("s(country)","s(studyID)"), 
                       se.fit=T)
plot(x_poedema, 100*inv.logit(predspoedema$fit), 
     panel.first = grid(), ylab='mortality (%)', xlab='', 
     type='p', lwd=3, ylim = c(0,24), xaxt='n',
     pch=19, xlim = c(-.25,1.25))
axis(1, at=x_poedema, labels = c('No\nPulmonary\nOedema','Pulmonary\nOedema'),
     tick = F,line = 1.5,cex.axis=.9)
ys = 100*inv.logit(c(predspoedema$fit + 1.96*predspoedema$se.fit,
                     (predspoedema$fit - 1.96*predspoedema$se.fit)))
arrows(x0 = x_poedema[1],y0 = ys[1], x1 = x_poedema[1], y1 = ys[3],angle = 90,
       lwd=6,code = 3,length=0,col = adjustcolor('blue',alpha.f = .3))
arrows(x0 = x_poedema[2],y0 = ys[2], x1 = x_poedema[2], y1 = ys[4],angle = 90,
       lwd=6,code = 3,length=0,col = adjustcolor('blue',alpha.f = .3))

#************** Shock
predsshock = predict(modshock, 
                     newdata = data.frame(shock=x_shock, 
                                          studyID='AQ', 
                                          country='Thailand'), 
                     exclude = c("s(country)","s(studyID)"), 
                     se.fit=T)
plot(x_shock, 100*inv.logit(predsshock$fit), 
     panel.first = grid(), ylab='mortality (%)', xlab='', 
     type='p', lwd=3, ylim = c(0,32), xaxt='n',
     pch=19, xlim = c(-.25,1.25))
axis(1, at=x_shock, labels = c('No shock','Shock'),
     tick = F,line = 0.5,cex.axis=.9)
ys = 100*inv.logit(c(predsshock$fit + 1.96*predsshock$se.fit,
                     (predsshock$fit - 1.96*predsshock$se.fit)))
arrows(x0 = x_shock[1],y0 = ys[1], x1 = x_shock[1], y1 = ys[3],angle = 90,
       lwd=6,code = 3,length=0,col = adjustcolor('blue',alpha.f = .3))
arrows(x0 = x_shock[2],y0 = ys[2], x1 = x_shock[2], y1 = ys[4],angle = 90,
       lwd=6,code = 3,length=0,col = adjustcolor('blue',alpha.f = .3))
```

Compute risk ratios with confidence intervals for the binary predictors:
```{r}
writeLines('\nComa risk ratio calculations:\n')
riskratio(x = table(Leg_data_complete$coma,Leg_data_complete$outcome),method = 'wald')
writeLines('\nShock risk ratio calculations:\n')
riskratio(x = table(Leg_data_complete$shock,Leg_data_complete$outcome),method = 'wald')
writeLines('\nConvulsions risk ratio calculations:\n')
riskratio(x = table(Leg_data_complete$convulsions,Leg_data_complete$outcome),method = 'wald')
writeLines('\nPulmonary oedema risk ratio calculations:\n')
riskratio(x = table(Leg_data_complete$poedema,Leg_data_complete$outcome),method = 'wald')
```


## Haematocrit distributions

```{r haematocrit_distributions}
par(las=1, bty='n')
f=ecdf(m$HCT[m$continent=="Africa"])
f2=ecdf(m$HCT[m$continent=="Asia"])
plot(5:50, 100*f(5:50), xlim=c(5,50), main='', lwd=2,xlab='Haematocrit (%)',
     xaxt='n',ylab='Cumulative proportion (%)',type='l',col='blue')
lines(5:50, 100*f2(5:50), col='green',lwd=2)
axis(1, at = c(5,10,15,20,25,35,45))

f_inverse = approxfun(x = f(1:50),y = 1:50)
f2_inverse = approxfun(x = f2(1:50),y = 1:50)

lines(c(-10, f2_inverse(.1)), c(10,10), lty=2)
lines(c(-10, f2_inverse(.5)), c(50,50), lty=2)
lines(c(f2_inverse(.5),f2_inverse(.5)), c(-10,50), lty=2)
lines(c(f_inverse(.5),f_inverse(.5)), c(-10,50), lty=2)
lines(c(f2_inverse(.1),f2_inverse(.1)), c(-10,10), lty=2)
lines(c(f_inverse(.1),f_inverse(.1)), c(-10,10), lty=2)

legend('bottomright',col=c('green','blue'),inset = 0.01,
       legend = c('Asian data','African data'),lwd=3)
```


# Sensitivity Analysis

## Only using complete records

```{r}
vars_interest = c('outcome','HCT','LPAR_pct','coma','convulsions','poedema',
                  'BUN','BD','shock','hypoglycaemia',
                  'studyID','country')
complete_cases = apply(m[,vars_interest], 1, function(x) sum(is.na(x))) == 0
Complete_Leg_data = m[complete_cases,] 
Complete_Leg_data$studyID = as.factor(as.character(Complete_Leg_data$studyID))
```

```{r}
DAG_fmla = "outcome ~ HCT + LPAR_pct + coma + convulsions + poedema + 
              log2(BUN) + BD + shock + hypoglycaemia + 
              drug_AS + (1 | studyID) + (1 | country)"
# fit the model to each dataset
mod1 = glmer(formula = DAG_fmla, data = Complete_Leg_data, 
             family=binomial) 
summary(mod1)
```

```{r}
# The scalar multiples to put the AORs on the correct scales
# Compute 95% CIs
FixedEffs = as.data.frame(summary(mod1)$coefficients)
FixedEffs$Scalar_f = c(1,10, 1, 1, 1, 1,log2(3), 10, 1, 1,1)
Results = data.frame(lowerCI = exp(FixedEffs$Scalar_f*(FixedEffs$Estimate -
                                                         1.96*FixedEffs$`Std. Error`)),
                     mean = exp(FixedEffs$Scalar_f*(FixedEffs$Estimate)),
                     upperCI = exp(FixedEffs$Scalar_f*(FixedEffs$Estimate +
                                                         1.96*FixedEffs$`Std. Error`)))
rownames(Results) = rownames(FixedEffs)
```

Make the 'forest' plot:
```{r ForestPlot_CompleteData}
plotting_ind = rownames(Results) %in% c('BD','coma','convulsions1','LPAR_pct',
                                        'drug_AS','HCT','log2(BUN)','poedema1','shock1')
Results = Results[plotting_ind,]
Results$Y_Labels = c('-10 % points\nabsolute haematocrit\non admission',
                     'Tenfold increase\nin parasitaemia',
                     'Coma\non admission',
                     'Seizures\non admission',
                     'Pulmonary\nOedema\non admission',
                     '3 fold increase\nin blood urea\nnitrogen (mmol/L)',
                     '+10 mEq/L\nbase deficit',
                     'Shock\non admission',
                     'Artemisinin drug\nversus\nnon Artemisinin drug'
)

Results['HCT',1:3] = 1/Results['HCT',1:3]
x_ind = sort.int(Results$mean, index.return = T)$ix
Results = Results[x_ind,]
par(bty='n', las=1, mar = c(3,9,2,2))

print(Results)

xlims = c(0.5, 8.5)
plot(NA,NA, xlim= log2(xlims), ylim = c(0,1),xaxt='n',
     xlab='', ylab='', yaxt='n')
axis(1, at = log2(c(0.5,1, 2,4)), labels = c(0.5,1, 2,4))
abline(v=0, lty=2, lwd=3, col='red')
yindex =1
ypos = seq(0,1,length.out = sum(plotting_ind))


for(i in 1:nrow(Results)){
  arrows(log2(Results[i,'lowerCI']),ypos[yindex],
         log2(Results[i,'upperCI']),ypos[yindex],
         length=0.0, angle=90, code=3, 
         col = 'black',lwd=3)
  points(log2(Results[i,'mean']),ypos[yindex],pch=18,cex=2)
  yindex=yindex+1
  
}
abline(h=ypos, lty=3)
axis(side = 2, at = ypos, labels = Results$Y_Labels,tick=FALSE)
mtext(side=1, line = 2, text = 'Adjusted odds ratio')
mtext(side = 3, line = 1, text = 'Increased survival',adj = 0)
mtext(side = 3, line = 1, text = 'Decreased survival',adj = 1)
```

# Trim very low haematocrits

```{r}
for (i in 1:length(SM_Impute_List)){
  SM_Impute_List[[i]] = filter(SM_Impute_List[[i]], HCT > 9)
}
```

```{r deathmodel_glm, include=FALSE, echo=TRUE}
if(RUN_MODELS){
  # the model formula
  DAG_fmla = "outcome ~ HCT + LPAR_pct + coma + convulsions + poedema + 
                        log2(BUN) + BD + shock + hypoglycaemia + 
                        drug_AS + (1 | continent/country)"
  # fit the model to each dataset
  modList= glmerModList(DAG_fmla, data = SM_Impute_List,
                        family=binomial,parallel = T) 
  # save the output
  save(modList, file = 'RData/Models_List_Remove_Hb_less_10.RData')
} else {
  load('RData/Models_List_Remove_Hb_less_10.RData')
}
```


```{r}
# extract the fixed and random effects from all the model fits
# These functions then compute the overall estimates
FixedEffs = modelFixedEff(modList)
RandEffs = modelRandEffStats(modList)
print(FixedEffs)
print(RandEffs)


FixedEffs$Scalar_f = as.numeric(mapvalues(x = FixedEffs$term, 
                                          from = c("(Intercept)",
                                                   "BD",
                                                   "coma",
                                                   "convulsions1",
                                                   "drug_AS",
                                                   "HCT",
                                                   "hypoglycaemia",
                                                   "log2(BUN)",
                                                   "LPAR_pct",
                                                   "shock1",
                                                   "poedema1"),
                                          to = as.numeric(c(1, 7, 1, 1, 1, 10, 1,2,log10(6),1, 1))))
FixedEffs$std.error=as.numeric(FixedEffs$std.error)
FixedEffs$estimate=as.numeric(FixedEffs$estimate)

# Compute 95% CIs
Results = data.frame(lowerCI = exp(FixedEffs$Scalar_f*(FixedEffs$estimate -
                                                         1.96*FixedEffs$std.error)),
                     mean = exp(FixedEffs$Scalar_f*(FixedEffs$estimate)),
                     upperCI = exp(FixedEffs$Scalar_f*(FixedEffs$estimate +
                                                         1.96*FixedEffs$std.error)))
rownames(Results) = FixedEffs$term
```

Aggregate results for plotting:
```{r}
writeLines(sprintf('Standard deviation of base deficit in patient population is %s',
                   round(sd(m$BD,na.rm=T))))
writeLines(sprintf('Standard deviation of blood urea nitrogen in patient population is %s',
                   round(2^sd(log2(m$BUN),na.rm=T))))
writeLines(sprintf('Standard deviation of parasitised RBCs in patient population is %s',
                   round(10^sd(m$LPAR_pct,na.rm=T))))
writeLines(sprintf('Standard deviation of haematocrits in patient population is %s',
                   round(sd(m$HCT,na.rm=T))))
```



Make the 'forest' plot:
```{r ForestPlot_SM_trimmedVerySMA}
Results = Results[rownames(Results) %in% c('BD','coma','convulsions1','drug_AS','HCT',
                                           'log2(BUN)','poedema1','LPAR_pct','shock1'),]
Results$Names =mapvalues(rownames(Results),
                         from = c("drug_AS",
                                  "HCT",
                                  "LPAR_pct",
                                  "convulsions1",
                                  "shock1",
                                  "poedema1",
                                  "log2(BUN)",
                                  "BD", 
                                  "coma"),
                         to = c('Artemisinin drug\nversus\nnon Artemisinin drug',
                                '-10 % points\nabsolute haematocrit\non admission',
                                'Six fold increase\n in parasitised\nred blood cells',
                                'Seizures\non admission',
                                'Shock\non admission',
                                'Pulmonary\noedema\non admission',
                                'Two fold increase\nin blood urea\nnitrogen (mmol/L)',
                                '+7 mEq/L\nbase deficit',
                                'Coma\non admission'))
Results['HCT',c("lowerCI","mean","upperCI")] = 1/Results['HCT',c("lowerCI","mean","upperCI")]
Results = arrange(Results, mean)
print(cbind(Results[,'Names'],round(Results[,c("lowerCI","mean","upperCI")],2)))

par(bty='n', las=1, mar = c(4,9,2,2))

xlims = c(0.5, 4.5)
plot(NA,NA, xlim= log2(xlims), ylim = c(0,1),xaxt='n',
     xlab='', ylab='', yaxt='n')
abline(v= seq(-1,3,by=1),col = "lightgray", lty = "dotted",lwd = par("lwd"))
axis(1, at = log2(c(0.5,1, 2,4)), labels = c(0.5,1, 2,4))
abline(v=0, lty=2, lwd=3, col='red')
yindex =1
ypos = seq(0,1,length.out = nrow(Results))


for(i in 1:nrow(Results)){
  arrows(log2(Results[i,'lowerCI']),ypos[yindex],
         log2(Results[i,'upperCI']),ypos[yindex],
         length=0.0, angle=90, code=3, 
         col = 'black',lwd=3)
  points(log2(Results[i,'mean']),ypos[yindex],pch=18,cex=2)
  yindex=yindex+1
  
}
abline(h=ypos, lty=3)
axis(side = 2, at = ypos, labels = Results$Names,tick=FALSE)
mtext(side=1, line = 2.5, text = 'Adjusted odds ratio')
mtext(side = 3, line = 1, text = 'Increased survival',adj = 0)
mtext(side = 3, line = 1, text = 'Decreased survival',adj = 1)
```

## The effect of Transfusion: AAV, AQ and AQUAMAT

For these three studies we have detailed transfusion data, with over half of patients being transfused:
```{r}
table(m$transfusion,m$studyID,useNA = 'ifany')
```

Select data from AQUAMAT, AAV and AQ:
```{r}
load('RData/Multiple_Imputed_Datasets.RData')
for(i in 1:length(SM_Impute_List)){
  SM_Impute_List[[i]] = filter(SM_Impute_List[[i]], studyID%in%c('AQUAMAT','AAV','AQ'))
}
```

### The naive analysis: no stratification by time of death


```{r}
DAG_fmla = "outcome ~ HCT + LPAR_pct + coma + convulsions + poedema + 
              log2(BUN) + BD + shock + hypoglycaemia + transfusion +
              drug_AS + (1 | country)"
if(RUN_MODELS){
  # fit the model to each dataset
  modList = glmerModList(DAG_fmla, data = SM_Impute_List,
                         family=binomial,parallel = T) 
  # save the output
  save(modList, file = 'RData/ModelList_transfusion_Naive.RData')
} else {
  load('RData/ModelList_transfusion_Naive.RData')
}

FixedEffs = modelFixedEff(modList)
print(FixedEffs)

FixedEffs$Scalar_f = as.numeric(mapvalues(x = FixedEffs$term, 
                                          from = c("(Intercept)", "BD","coma", "convulsions1",
                                                   "drug_AS","HCT","hypoglycaemia", "log2(BUN)",
                                                   "LPAR_pct","shock1","poedema1", "transfusion"),
                                          to = as.numeric(c(1, 7, 1, 1, 1, 
                                                            10, 1,2,log10(6),1, 1,1))))
FixedEffs$std.error=as.numeric(FixedEffs$std.error)
FixedEffs$estimate=as.numeric(FixedEffs$estimate)

# Compute 95% CIs
Results = data.frame(lowerCI = exp(FixedEffs$Scalar_f*(FixedEffs$estimate -
                                                         1.96*FixedEffs$std.error)),
                     mean = exp(FixedEffs$Scalar_f*(FixedEffs$estimate)),
                     upperCI = exp(FixedEffs$Scalar_f*(FixedEffs$estimate +
                                                         1.96*FixedEffs$std.error)))
rownames(Results) = FixedEffs$term
Results['HCT',1:3] = 1/Results['HCT',1:3]
Results$Names =mapvalues(rownames(Results),
                         from = c("drug_AS","HCT","LPAR_pct","convulsions1","shock1",
                                  "poedema1","log2(BUN)","BD", "coma","transfusion"),
                         to = c('Artemisinin drug\nversus\nnon Artemisinin drug',
                                '-10 % points\nabsolute haematocrit\non admission',
                                'Six fold increase\n in parasitised\nred blood cells',
                                'Seizures\non admission',
                                'Shock\non admission',
                                'Pulmonary\noedema\non admission',
                                'Two fold increase\nin blood urea\nnitrogen (mmol/L)',
                                '+7 mEq/L\nbase deficit',
                                'Coma\non admission','Transfusion'))
```


Make the 'forest' plot:
```{r ForestPlot_Transfusion_Naive}
plotting_ind = rownames(Results) %in% c('BD','coma','convulsions1','LPAR_pct','transfusion',
                                        'drug_AS','HCT','log2(BUN)','poedema1','shock1')
Results = Results[plotting_ind,]



x_ind = sort.int(Results$mean, index.return = T)$ix
Results = Results[x_ind,]
par(bty='n', las=1, mar = c(3,9,2,2))

print(Results)

xlims = c(0.5, 8.5)
plot(NA,NA, xlim= log2(xlims), ylim = c(0,1),xaxt='n',
     xlab='', ylab='', yaxt='n')
axis(1, at = log2(c(0.5,1, 2,4)), labels = c(0.5,1, 2,4))
abline(v=0, lty=2, lwd=3, col='red')
yindex =1
ypos = seq(0,1,length.out = sum(plotting_ind))


for(i in 1:nrow(Results)){
  arrows(log2(Results[i,'lowerCI']),ypos[yindex],
         log2(Results[i,'upperCI']),ypos[yindex],
         length=0.0, angle=90, code=3, 
         col = 'black',lwd=3)
  points(log2(Results[i,'mean']),ypos[yindex],pch=18,cex=2)
  yindex=yindex+1
  
}
abline(h=ypos, lty=3)
axis(side = 2, at = ypos, labels = Results$Names,tick=FALSE)
mtext(side=1, line = 2, text = 'Adjusted odds ratio')
mtext(side = 3, line = 1, text = 'Increased survival',adj = 0)
mtext(side = 3, line = 1, text = 'Decreased survival',adj = 1)
```

We see that transfusion is associated with a lower odds of death.
However, this is a biased estimate as transfusion is a collider (see paper).

### Stratification by time to death

We stratify by time to death, to remove collider bias.
Select the patients who survive past 4 hours
```{r}
#### We're just looking at the those who didn't die before 4 hours post admission
m$Outcome4hours = 0
m$Outcome4hours[!is.na(m$Timetodeathhrs) & m$Timetodeathhrs < 4] = 1
# We only look at individuals who survive past 4 hours
m = filter(m, Outcome4hours==0)

Transfusion_Data = m[m$studyID%in%c('AQUAMAT','AAV','AQ'),]
round(100*table(Transfusion_Data$transfusion)/nrow(Transfusion_Data),1)
unique_ids_Transfusion = Transfusion_Data$Unique_ID
```

We load the imputed datasets and select patients whose transfusion information was recorded (AQ,AAV and AQUAMAT), and who did not die in the first 4 hours
```{r}
for(i in 1:length(SM_Impute_List)){
  ind = SM_Impute_List[[i]]$Unique_ID %in% unique_ids_Transfusion
  SM_Impute_List[[i]] = SM_Impute_List[[i]][ind,]
}

SM_Impute_List = SM_Impute_List[1:50]
```

Fit the model to these data:
```{r, include=FALSE}
if(RUN_MODELS){
  DAG_fmla = "outcome ~ HCT + LPAR_pct + coma + convulsions + 
              log2(BUN) + BD + shock + hypoglycaemia + transfusion +
              drug_AS + (1 | country)"
  # fit the model to each dataset
  modList = glmerModList(DAG_fmla, data = SM_Impute_List,
                         family=binomial,parallel = T) 
  # save the output
  save(modList, file = 'RData/Models_List_TransfusionMediation.RData')
} else {
  load('RData/Models_List_TransfusionMediation.RData')
}
```

Compute the overall parameter estimates:
```{r}
# extract the fixed effects from all the model fits
# These functions then compute the overall estimates
FixedEffs = modelFixedEff(modList)
print(FixedEffs)
```

```{r}
# The scalar multiples to put the AORs on the correct scales
# Compute 95% CIs
FixedEffs$Scalar_f = as.numeric(mapvalues(FixedEffs$term, 
                                          from = c('(Intercept)',
                                                   "hypoglycaemia",
                                                   'BD',
                                                   'coma',
                                                   'convulsions1',
                                                   'LPAR_pct',
                                                   'transfusion',
                                                   'drug_AS',
                                                   'HCT',
                                                   'log2(BUN)',
                                                   'poedema1',
                                                   'shock1'),
                                          to = c(1,1,10, 1, 1, 1, 1,1,10,log2(3),1,1)))
Results = data.frame(lowerCI = exp(FixedEffs$Scalar_f*(FixedEffs$estimate -
                                                         1.96*FixedEffs$`std.error`)),
                     mean = exp(FixedEffs$Scalar_f*(FixedEffs$estimate)),
                     upperCI = exp(FixedEffs$Scalar_f*(FixedEffs$estimate +
                                                         1.96*FixedEffs$`std.error`)))
Results$term = FixedEffs$term
plotting_ind = Results$term %in% c('BD','coma','convulsions1','LPAR_pct','transfusion',
                                   'drug_AS','HCT','log2(BUN)','poedema1','shock1')
Results = Results[plotting_ind,]
Results$Y_Labels = mapvalues(Results$term, 
                             from = c('BD',
                                      'coma',
                                      'convulsions1',
                                      'LPAR_pct',
                                      'transfusion',
                                      'drug_AS',
                                      'HCT',
                                      'log2(BUN)',
                                      'poedema1',
                                      'shock1'),
                             to = c('+10 mEq/L\nbase deficit',
                                    'Coma\non admission',
                                    'Seizures\non admission',
                                    'Tenfold increase\nin parasitaemia',
                                    'Transfusion',
                                    'Artemisinin drug\nversus\nnon Artemisinin drug',
                                    '-10 % points\nabsolute haematocrit\non admission',
                                    '3 fold increase\nin blood urea\nnitrogen (mmol/L)',
                                    'Pulmonary\nOedema\non admission',
                                    'Shock\non admission'))
```


Make the 'forest' plot:
```{r ForestPlot_Transfusion_TD_stratified}
Results[Results$term=='HCT',1:3] = 1/Results[Results$term=='HCT',1:3]
x_ind = sort.int(Results$mean, index.return = T)$ix
Results = Results[x_ind,]
par(bty='n', las=1, mar = c(3,9,2,2))

print(Results)

xlims = c(0.5, 4)
plot(NA,NA, xlim= log2(xlims), ylim = c(0,1),xaxt='n',
     xlab='', ylab='', yaxt='n')
axis(1, at = log2(c(0.5,1, 2,4)), labels = c(0.5,1, 2,4))
abline(v=0, lty=2, lwd=3, col='red')
yindex =1
ypos = seq(0,1,length.out = sum(plotting_ind))


for(i in 1:nrow(Results)){
  arrows(log2(Results[i,'lowerCI']),ypos[yindex],
         log2(Results[i,'upperCI']),ypos[yindex],
         length=0.0, angle=90, code=3, 
         col = 'black',lwd=3)
  points(log2(Results[i,'mean']),ypos[yindex],pch=18,cex=2)
  yindex=yindex+1
  
}
abline(h=ypos, lty=3)
axis(side = 2, at = ypos, labels = Results$Y_Labels,tick=FALSE)
mtext(side=1, line = 2, text = 'Adjusted odds ratio')
mtext(side = 3, line = 1, text = 'Increased survival',adj = 0)
mtext(side = 3, line = 1, text = 'Decreased survival',adj = 1)
```




