---
title: "Charactersing effect of anaemia on mortality in severe malaria"
output:
  html_document:
    keep_md: yes
    fig_caption: yes
---

# Preliminaries


```{r, echo=F}
knitr::opts_chunk$set(cache = TRUE, cache.comments = FALSE, 
                      echo = TRUE, include = TRUE, 
                      fig.width = 7, fig.height = 7,
                      fig.pos = 'H', 
                      dev = 'png', dpi = 300)
```

```{r, echo=FALSE, include=FALSE}
library(lme4)
require(RColorBrewer)
require(boot)
require(plyr)
require(dplyr)
library(mgcv) # For the GAM modelling
```



# Exploratory analysis


This looks at the severe malaria legacy dataset from MORU.

```{r, echo=FALSE}
load('RData/Data.RData')
ind_missing = apply(m[,c('AgeInYear','HCT','BUN','BD','LPAR_pct')], 1, function(x) sum(is.na(x))>0)
Leg_data_complete = m[!ind_missing,]
Leg_data_complete = Leg_data_complete[!is.infinite(Leg_data_complete$LPAR_pct),]
Leg_data_complete$studyID = as.factor(Leg_data_complete$studyID)# for plotting
```



First we look at the mortality rates across the different studies.

```{r, echo=FALSE}
for(s in unique(m$studyID)){
  print(paste(s, ', mortality of:', round(100*mean(m$outcome[m$studyID==s])),'%'))
}
```

We look at the quantiles of the ages in the different studies:
```{r, echo=FALSE}
for(s in unique(m$studyID)){
  print(paste0(s, ', ages:', round(quantile(m$AgeInYear[m$studyID==s], probs = c(0,.5,1))), collapse = ' '))
}

for(s in unique(m$studyID)){
  print(s)
  print(table(m$drug[m$studyID==s]))
}
```


Let's look at the linear associations between the key baseline variables. We use mixed effects linear models to estimate these associations (random effect terms for both country and study).

```{r , echo=FALSE}
mod_Age_HCT = gam(formula = HCT ~ s(AgeInYear)  + s(studyID, bs='re') + 
                    s(country, bs='re'), 
                  data = Leg_data_complete)
mod_HCT_BD = gam(formula = BD ~ s(HCT) + s(studyID, bs='re') + 
                   s(country, bs='re'), 
                 data = Leg_data_complete)
mod_BD_LPAR = gam(formula = BD ~ s(LPAR_pct) + s(studyID, bs='re') + 
                    s(country, bs='re'),
                  data =Leg_data_complete)
mod_BUN_BD = gam(formula = BD ~ s(log10(BUN)) + s(studyID, bs='re') + 
                   s(country, bs='re'),
                 data =Leg_data_complete)

```

```{r ExploratoryPlots, echo=FALSE}
mycols = brewer.pal(length(unique(m$studyID)), 'Set1')
Leg_data_complete$color = as.character(revalue(Leg_data_complete$studyID, 
                                               replace = c('Core Malaria'=mycols[1],
                                                           'AAV'=mycols[2],
                                                           'AQ'=mycols[3],
                                                           'SEAQUAMAT' = mycols[4],
                                                           'AQUAMAT' = mycols[5])))
par(las=1, mfrow=c(2,2), mar=c(4,4,1,1), bty='n')
## Base Excess and HCT
plot(jitter(Leg_data_complete$HCT,amount=1), 
     jitter(Leg_data_complete$BD), 
     col=Leg_data_complete$color, pch='*', xlab='Haematocrit (%)', 
     ylab='Base deficit (mEq/L)')

ys = predict(object = mod_HCT_BD, 
             newdata = data.frame(HCT=8:50,
                                  studyID='AQ',
                                  country='Vietnam'), 
             exclude = c("s(country)","s(studyID)"))
lines(8:50, ys, lwd=3, col='black')

## Parasitaemia and Anaemia
plot(jitter(Leg_data_complete$LPAR_pct,amount=1), 
     Leg_data_complete$BD, 
     col=Leg_data_complete$color, pch='*', 
     xlab='Log10 % parasitised RBCs', 
     ylab='Base deficit (mEq/L)')
ys = predict(object = mod_BD_LPAR, 
             newdata = data.frame(LPAR_pct=seq(-3,3,by=.1),
                                  studyID='AQ',
                                  country='Vietnam'), 
             exclude = c("s(country)","s(studyID)"))
lines(seq(-3,3,by=.1), ys, lwd=3, col='black')

## BUN and BD
plot(jitter(log10(Leg_data_complete$BUN),amount=.1), 
     jitter(Leg_data_complete$BD), 
     col=Leg_data_complete$color, pch='*', 
     xlab='Blood urea nitrogen (mmol/L)', 
     ylab='Base deficit (mEq/L)', xaxt='n')
axis(1, at=c(log10(2), 1, 2), labels = c(2,10,100))

summary(mod_BUN_BD)
ys = predict(object = mod_BUN_BD, 
             newdata = data.frame(BUN=2:166,
                                  studyID='AQ',
                                  country='Vietnam'), 
             exclude = c("s(country)","s(studyID)"))
lines(log10(2:166), ys, lwd=3, col='black')

## Parasitaemia and Anaemia
plot(jitter(Leg_data_complete$AgeInYear,amount=1), 
     Leg_data_complete$HCT, 
     col=Leg_data_complete$color, pch='*', 
     xlab='Age (years)', 
     ylab='Haematocrit (%)')

summary(mod_Age_HCT)
ys = predict(object = mod_Age_HCT, 
             newdata = data.frame(AgeInYear=0:80,
                                  studyID='AQ',
                                  country='Vietnam'), 
             exclude = c("s(country)","s(studyID)"))
lines(0:80, ys, lwd=3, col='black')
legend('topright',col=mycols,legend = c('Core Malaria',
                                        'AAV',
                                        'AQ',
                                        'SEAQUAMAT',
                                        'AQUAMAT'),pch='*')
```

Effect on survival 

```{r UnAdjusted}
modHCT=gam(outcome ~ s(HCT) + s(studyID, bs='re') + s(country, bs='re'),
           data = Leg_data_complete, family='binomial')

modcoma=gam(outcome ~ coma + s(studyID, bs='re') + s(country, bs='re'),
            data = Leg_data_complete, family='binomial')

modBD=gam(outcome ~ s(BD) + s(studyID, bs='re') + s(country, bs='re'),
          data = Leg_data_complete, family='binomial')

modpoedema=gam(outcome ~ poedema + s(studyID, bs='re') + s(country, bs='re'),
               data = Leg_data_complete, family='binomial')

modconv=gam(outcome ~ convulsions + s(studyID, bs='re') + s(country, bs='re'),
            data = Leg_data_complete, family='binomial')

modBUN=gam(outcome ~ s(log10(BUN)) + s(studyID, bs='re') + s(country, bs='re'),
           data = Leg_data_complete, family='binomial')
```

```{r UnadjustedPlots, echo=FALSE}
par(las=1, bty='n', mfrow=c(2,3))

Factors = c('HCT','coma','BD','poedema','convulsions','BUN')
x_hcts = seq(4,45, by=1)
x_comas = 0:1
x_conv = 0:1
x_poedema = 0:1
x_bd = seq(-22,35, by=2)
x_BUN = seq(2,150, length.out = 2)

#************** BD
predsBD = predict(modBD, newdata = data.frame(BD= x_bd, 
                                              studyID='AQ', 
                                              country='Thailand'), 
                  exclude = c("s(country)","s(studyID)"), 
                  se.fit=T)
plot(x_bd, 100*inv.logit(predsBD$fit), ylab='%', xlab='Base deficit (mEq/L)', 
     type='l', lwd=3,ylim = c(0,100))
polygon(c(x_bd, rev(x_bd)), 100*inv.logit(c(predsBD$fit + 1.96*predsBD$se.fit,
                                            rev(predsBD$fit - 1.96*predsBD$se.fit))),
        lty=2,lwd=2, border = NA, col = rgb(0, 0, 1,0.5))

#************** HCT
predsHCT = predict(modHCT, 
                   newdata = data.frame(HCT=x_hcts, 
                                        studyID='AQ', 
                                        country='Thailand'), 
                   exclude = c("s(country)","s(studyID)"), 
                   se.fit=T)
plot(x_hcts, 100*inv.logit(predsHCT$fit), ylab='%', xlab='Haematocrit (%)', 
     type='l', lwd=3,ylim = c(5,37))
polygon(c(x_hcts,rev(x_hcts)), 
        100*inv.logit(c(predsHCT$fit + 1.96*predsHCT$se.fit, 
                        rev(predsHCT$fit - 1.96*predsHCT$se.fit))),
        lty=2,lwd=2,border = NA, col = rgb(0, 0, 1,0.5))

#************** BUN
predsBUN = predict(modBUN, newdata = data.frame(BUN=x_BUN, 
                                                studyID='AQ', 
                                                country='Thailand'), 
                   exclude = c("s(country)","s(studyID)"), 
                   se.fit=T)
plot(log10(x_BUN), 100*inv.logit(predsBUN$fit), ylab='%', 
     xlab='Blood urea nitrogen (mmol/L)', 
     type='l', lwd=3,ylim = c(0,60), xaxt='n')
axis(1, at = log10(c(2,10,100)), labels = c(2,10,100))
polygon(c(log10(x_BUN),rev(log10(x_BUN))), 
        100*inv.logit(c(predsBUN$fit + 1.96*predsBUN$se.fit,
                        rev(predsBUN$fit - 1.96*predsBUN$se.fit))), 
        lty=2,lwd=2, border = NA, col = rgb(0, 0, 1,0.5))

#**************  Coma
predscoma = predict(modcoma, newdata = data.frame(coma=x_comas, 
                                                  studyID='AQ', 
                                                  country='Thailand'), 
                    exclude = c("s(country)","s(studyID)"), 
                    se.fit=T)
plot(x_comas, 100*inv.logit(predscoma$fit), ylab='%', xlab='', 
     type='l', lwd=3, ylim = c(0,26), xaxt='n')
axis(1, at=x_comas, labels = c('No Coma','Coma'))
polygon(c(x_comas,rev(x_comas)), 
        100*inv.logit(c(predscoma$fit + 1.96*predscoma$se.fit,
                        rev(predscoma$fit -1.96*predscoma$se.fit))),
        lty=2,lwd=2, border = NA, col = rgb(0, 0, 1,0.5))

#************** Convulsions
predsconv = predict(modconv, newdata = data.frame(convulsions=x_conv, 
                                                  studyID='AQ', 
                                                  country='Thailand'), 
                    exclude = c("s(country)","s(studyID)"), 
                    se.fit=T)
plot(x_conv, 100*inv.logit(predsconv$fit), ylab='%', xlab='', 
     type='l', lwd=3, ylim = c(0,30), xaxt='n')
axis(1, at=x_conv, labels = c('No convulsions','Convulsions'))
polygon(c(x_conv,rev(x_conv)), 
        100*inv.logit(c(predsconv$fit + 1.96*predsconv$se.fit,
                        rev(predsconv$fit -1.96*predsconv$se.fit))),
        lty=2,lwd=2, border = NA, col = rgb(0, 0, 1,0.5))

#************** Pulmonary Oedema
predspoedema = predict(modpoedema, 
                       newdata = data.frame(poedema=x_poedema, 
                                            studyID='AQ', 
                                            country='Thailand'), 
                       exclude = c("s(country)","s(studyID)"), 
                       se.fit=T)
plot(x_poedema, 100*inv.logit(predspoedema$fit), ylab='%', xlab='', 
     type='l', lwd=3, ylim = c(0,55), xaxt='n')
axis(1, at=x_poedema, labels = c('No PO','PO'))
polygon(c(x_poedema,rev(x_poedema)), 
        100*inv.logit(c(predspoedema$fit + 1.96*predspoedema$se.fit,
          rev(predspoedema$fit - 1.96*predspoedema$se.fit))),
        lty=2,lwd=2, border = NA, col = rgb(0, 0, 1,0.5))
```

# Sensitivity Analysis

## Imputation of missing variables

Quite a lot of the important covariates are missing in the older studies. We use linear regression to estimate these unknown variables. This section shows the results for single imputation - when fitting the final models we use multiple imputation.

* Mising base deficit is imputed using bicarbonate (if available) else using respiratory rate
* Missing Blood urea nitrogen is imputed using creatinine

Impute base deficit from bicarbonate
```{r}
BD_and_bicarbonate = !is.na(m$BD) & !is.na(m$bicarbonate)
print(paste('We have ', sum(BD_and_bicarbonate), 'observations for both bicarbonate and base deficit'))
mod_impute1 = lmer(BD ~ bicarbonate + (1 | studyID) + (1 | country), data= m[BD_and_bicarbonate,])
missing_BD = is.na(m$BD)
Available_Bicarbonate = !is.na(m$bicarbonate)
print(paste(sum(missing_BD & Available_Bicarbonate), 'observations will now be imputed'))
# impute with model
m$BD[missing_BD & Available_Bicarbonate] = predict(mod_impute1,newdata=m[missing_BD & Available_Bicarbonate,], re.form=NA)
```

Impute base deficit from lactate
```{r}
BD_and_lactate = !is.na(m$BD) & !is.na(m$lactate)
print(paste('We have ', sum(BD_and_lactate), 'observations for both lactate and base deficit'))
if(length(unique(m$studyID[BD_and_lactate]))==1){
  mod_impute2 = lm(BD ~ lactate, data= m[BD_and_lactate,])
} else {
  mod_impute2 = lmer(BD ~ lactate + (1 | studyID), data= m[BD_and_lactate,])
}
missing_BD = is.na(m$BD)
Available_Lactate = !is.na(m$lactate)
print(paste(sum(missing_BD & Available_Lactate), 'observations will now be imputed'))
# impute with model
m$BD[missing_BD & Available_Lactate] = predict(mod_impute2,newdata=m[missing_BD & Available_Lactate,], re.form=NA)
```

Impute base deficit from respiratory rate
```{r}
BD_and_rr = !is.na(m$BD) & !is.na(m$rr)
print(paste('We have ', sum(BD_and_rr), 'observations for both resp rate and base deficit'))
mod_impute3 = lmer(BD ~ rr + (1 | studyID), data= m[BD_and_rr,])
missing_BD = is.na(m$BD)
Available_rr = !is.na(m$rr)
print(paste(sum(missing_BD & Available_rr), 'observations will now be imputed'))
m$BD[missing_BD & Available_rr] = predict(mod_impute3,newdata=m[missing_BD & Available_rr,], re.form=NA)
```


Impute blood urea nitrogen from creatinine:
```{r}
BUN_and_cr = !is.na(m$BUN) & !is.na(m$creatinine)
print(paste('We have ', sum(BUN_and_cr), 'observations for both blood urea nitrogen and creatinine'))
mod_impute4 = lmer(BUN ~ creatinine + (1 | studyID), data= m[BUN_and_cr,])
missing_BUN = is.na(m$BUN)
Available_cr = !is.na(m$creatinine)
print(paste(sum(missing_BUN & Available_cr), 'observations will now be imputed'))
m$BUN[missing_BUN & Available_cr] = predict(mod_impute4,newdata=m[missing_BUN & Available_cr,], re.form=NA)
```

Resulting data we can now use:
The contributions of the different studies:
```{r}
vars_interest = c('outcome','HCT','LPAR_pct','coma','convulsions','poedema',
                  'BUN','BD','shock','hypoglycaemia',
                  'studyID','country')
complete_cases = apply(m[,vars_interest], 1, function(x) sum(is.na(x))) == 0
Complete_Leg_data = m[complete_cases,] # for the model fitting
Complete_Leg_data$studyID = as.factor(as.character(Complete_Leg_data$studyID))
```



```{r}
mod_full_GAM = gam(outcome ~ s(HCT,AgeInYear) + LPAR_pct  + coma + convulsions +
                     poedema + log10(BUN) + BD + drug_AS + 
                     s(studyID, bs='re') + s(country, bs='re'),
                   data=Complete_Leg_data, family=binomial)
summary(mod_full_GAM)
```

