---
title: "Charactersing effect of anaemia on mortality in severe malaria"
output:
  html_document:
    keep_md: yes
    fig_caption: yes
---

# Preliminaries


```{r, echo=F}
knitr::opts_chunk$set(cache = TRUE, cache.comments = FALSE, 
                      echo = TRUE, include = TRUE, 
                      fig.width = 7, fig.height = 7,
                      fig.pos = 'H', 
                      dev = 'png', dpi = 300)

RUN_MODELS = F
```

```{r, echo=FALSE, include=FALSE}
library(lme4)
require(RColorBrewer)
library(merTools)
require(boot)
require(plyr)
require(dplyr)
library(mgcv) # For the GAM modelling
```



# Exploratory analysis


This looks at the severe malaria legacy dataset from MORU.

```{r, echo=FALSE}
load('RData/Data.RData')
m$coma[is.na(m$coma) & m$studyID=='AQ'] = 0

m$drug_AS = 0
m$drug_AS[m$drug_class=='artemisinin']=1
m$LPAR_pct[is.infinite(m$LPAR_pct)] = 0

ind_missing = apply(m[,c('AgeInYear','HCT','BUN','BD','LPAR_pct')], 1, function(x) sum(is.na(x))>0)
Leg_data_complete = m[!ind_missing,]
Leg_data_complete = Leg_data_complete[!is.infinite(Leg_data_complete$LPAR_pct),]
Leg_data_complete$studyID = as.factor(Leg_data_complete$studyID)# for plotting
```



First we look at the mortality rates across the different studies.

```{r, echo=FALSE}
for(s in unique(m$studyID)){
  print(paste(s, ', mortality of:', round(100*mean(m$outcome[m$studyID==s])),'%'))
}
```

We look at the quantiles of the ages in the different studies:
```{r, echo=FALSE}
for(s in unique(m$studyID)){
  print(paste0(s, ', ages:', round(quantile(m$AgeInYear[m$studyID==s], probs = c(0,.5,1))), collapse = ' '))
}

for(s in unique(m$studyID)){
  print(s)
  print(table(m$drug[m$studyID==s]))
}
```


Let's look at the linear associations between the key baseline variables. We use mixed effects linear models to estimate these associations (random effect terms for both country and study).

```{r , echo=FALSE}
mod_Age_HCT = gam(formula = HCT ~ s(AgeInYear)  + s(studyID, bs='re') + 
                    s(country, bs='re'), 
                  data = Leg_data_complete)
mod_HCT_BD = gam(formula = BD ~ s(HCT) + s(studyID, bs='re') + 
                   s(country, bs='re'), 
                 data = Leg_data_complete)
mod_BD_LPAR = gam(formula = BD ~ s(LPAR_pct) + s(studyID, bs='re') + 
                    s(country, bs='re'),
                  data =Leg_data_complete)
mod_BUN_BD = gam(formula = BD ~ s(log10(BUN)) + s(studyID, bs='re') + 
                   s(country, bs='re'),
                 data =Leg_data_complete)
```

```{r ExploratoryPlots, echo=FALSE}
mycols = brewer.pal(length(unique(m$studyID)), 'Set1')
Leg_data_complete$color = as.character(revalue(Leg_data_complete$studyID, 
                                               replace = c('Core Malaria'=mycols[1],
                                                           'AAV'=mycols[2],
                                                           'AQ'=mycols[3],
                                                           'SEAQUAMAT' = mycols[4],
                                                           'AQUAMAT' = mycols[5])))
par(las=1, mfrow=c(2,2), mar=c(4,4,1,1), bty='n')
## Base Excess and HCT
plot(jitter(Leg_data_complete$HCT,amount = .4), 
     jitter(Leg_data_complete$BD, amount = .4), ylim=c(-6,28),
     col=Leg_data_complete$color, pch='*', xlab='', 
     ylab='Base deficit (mEq/L)', xlim=c(8,47), panel.first = grid())
mtext(text = 'Haematocrit (%)',side = 1,line = 2.5)
ys = predict(object = mod_HCT_BD, 
             newdata = data.frame(HCT=8:50,
                                  studyID='AQ',
                                  country='Vietnam'), 
             exclude = c("s(country)","s(studyID)"))
lines(8:50, ys, lwd=3, col='black')

## Parasitaemia and Anaemia
plot(Leg_data_complete$LPAR_pct, 
     jitter(Leg_data_complete$BD,amount = .4), 
     xlim = c(-2.418006, 1.562435),ylim=c(-6,28),
     col=Leg_data_complete$color, pch='*', 
     xlab='', panel.first = grid(),
     ylab='Base deficit (mEq/L)', xaxt='n')
axis(1, at = log10(c(30,10,1,.1,.01)), labels = c(30,10,1,.1,.01))
mtext(text = 'Parasitised RBCs (%)',side = 1,line = 2.5)
ys = predict(object = mod_BD_LPAR, 
             newdata = data.frame(LPAR_pct=seq(-3,log10(50),by=.1),
                                  studyID='AQ',
                                  country='Vietnam'), 
             exclude = c("s(country)","s(studyID)"))
lines(seq(-3,log10(50),by=.1), ys, lwd=3, col='black')

## BUN and BD
plot(jitter(log10(Leg_data_complete$BUN),amount = .4),
     jitter(Leg_data_complete$BD, amount = .4), ylim=c(-6,28),
     col=Leg_data_complete$color, pch='*', panel.first = grid(),
     xlab='',xlim = c(0.4771213,2.1461280),
     ylab='Base deficit (mEq/L)', xaxt='n')
mtext(text = 'Blood urea nitrogen (mmol/L)',side = 1,line = 2.5)
axis(1, at=c(log10(3), log10(5), 1, log10(30),2), labels = c(3,5,10,30,100))

summary(mod_BUN_BD)
ys = predict(object = mod_BUN_BD, 
             newdata = data.frame(BUN=2:166,
                                  studyID='AQ',
                                  country='Vietnam'), 
             exclude = c("s(country)","s(studyID)"))
lines(log10(2:166), ys, lwd=3, col='black')

## Parasitaemia and Anaemia
plot(jitter(Leg_data_complete$AgeInYear,amount=.4), 
     jitter(Leg_data_complete$HCT,amount = .4), 
     col=Leg_data_complete$color, pch='*', 
     xlab='', ylim=c(8,47), panel.first = grid(),
     ylab='Haematocrit (%)', xlim = c(0,65))
mtext(text = 'Age (years)',side = 1,line = 2.5)
summary(mod_Age_HCT)
ys = predict(object = mod_Age_HCT, 
             newdata = data.frame(AgeInYear=0:80,
                                  studyID='AQ',
                                  country='Vietnam'), 
             exclude = c("s(country)","s(studyID)"))
lines(0:80, ys, lwd=3, col='black')
legend('bottomright',col=mycols,inset = 0.01,
       legend = c('Core Malaria',
                  'AAV',
                  'AQ',
                  'SEAQUAMAT',
                  'AQUAMAT'),pch='*')
```

Effect on survival 

```{r UnAdjusted}
modHCT=gam(outcome ~ s(HCT) + s(studyID, bs='re') + s(country, bs='re'),
           data = Leg_data_complete, family='binomial')

modcoma=gam(outcome ~ coma + s(studyID, bs='re') + s(country, bs='re'),
            data = Leg_data_complete, family='binomial')

modBD=gam(outcome ~ s(BD) + s(studyID, bs='re') + s(country, bs='re'),
          data = Leg_data_complete, family='binomial')

modpoedema=gam(outcome ~ poedema + s(studyID, bs='re') + s(country, bs='re'),
               data = Leg_data_complete, family='binomial')

modconv=gam(outcome ~ convulsions + s(studyID, bs='re') + s(country, bs='re'),
            data = Leg_data_complete, family='binomial')

modshock=gam(outcome ~ shock + s(studyID, bs='re') + s(country, bs='re'),
            data = Leg_data_complete, family='binomial')

modBUN=gam(outcome ~ s(log10(BUN)) + s(studyID, bs='re') + s(country, bs='re'),
           data = Leg_data_complete, family='binomial')

modLPAR_pct=gam(outcome ~ s(LPAR_pct) + s(studyID, bs='re') + s(country, bs='re'),
           data = Leg_data_complete, family='binomial')
```

```{r UnadjustedPlots, echo=FALSE}
par(las=1, bty='n', mfrow=c(2,4))

Factors = c('HCT','coma','BD','poedema','convulsions','BUN')
x_hcts = seq(10,45, by=1)
x_comas = 0:1
x_shock = 0:1
x_conv = 0:1
x_poedema = 0:1
x_bd = seq(-5,35, by=2)
x_BUN = 2:150
x_parasites = seq(-2,log10(30),by=.001)

#************** BD
predsBD = predict(modBD, newdata = data.frame(BD= x_bd, 
                                              studyID='AQ', 
                                              country='Thailand'), 
                  exclude = c("s(country)","s(studyID)"), 
                  se.fit=T)
plot(x_bd, 100*inv.logit(predsBD$fit), ylab='mortality (%)', 
     panel.first = grid(), xlab='', 
     type='l', lwd=3,ylim = c(0,80))
mtext(text = 'Base deficit (mEq/L)',side = 1,line = 2.5, cex=.7)
polygon(c(x_bd, rev(x_bd)), 100*inv.logit(c(predsBD$fit + 1.96*predsBD$se.fit,
                                            rev(predsBD$fit - 1.96*predsBD$se.fit))),
        lty=2,lwd=2, border = NA, col = rgb(0, 0, 1,0.5))

#************** HCT
predsHCT = predict(modHCT, 
                   newdata = data.frame(HCT=x_hcts, 
                                        studyID='AQ', 
                                        country='Thailand'), 
                   exclude = c("s(country)","s(studyID)"), 
                   se.fit=T)
plot(x_hcts, 100*inv.logit(predsHCT$fit), ylab='mortality (%)', 
     panel.first = grid(), xlab='', 
     type='l', lwd=3,ylim = c(0,25))
mtext(text = 'Haematocrit (%)',side = 1,line = 2.5,cex=.7)
polygon(c(x_hcts,rev(x_hcts)), 
        100*inv.logit(c(predsHCT$fit + 1.96*predsHCT$se.fit, 
                        rev(predsHCT$fit - 1.96*predsHCT$se.fit))),
        lty=2,lwd=2,border = NA, col = rgb(0, 0, 1,0.5))

#************** BUN
predsBUN = predict(modBUN, newdata = data.frame(BUN=x_BUN, 
                                                studyID='AQ', 
                                                country='Thailand'), 
                   exclude = c("s(country)","s(studyID)"), 
                   se.fit=T)
plot(log10(x_BUN), 100*inv.logit(predsBUN$fit), ylab='mortality (%)', 
     xlab='', panel.first = grid(), 
     type='l', lwd=3,ylim = c(0,60), xaxt='n')
mtext(text = 'Blood urea nitrogen (mmol/L)',side = 1,line = 2.5,cex=.7)
axis(1, at = log10(c(2,10,100)), labels = c(2,10,100))
polygon(c(log10(x_BUN),rev(log10(x_BUN))), 
        100*inv.logit(c(predsBUN$fit + 1.96*predsBUN$se.fit,
                        rev(predsBUN$fit - 1.96*predsBUN$se.fit))), 
        lty=2,lwd=2, border = NA, col = rgb(0, 0, 1,0.5))

#************** Log10 parasitaemia
predspara_PCT = predict(modLPAR_pct, newdata = data.frame(LPAR_pct=x_parasites, 
                                                studyID='AQ', 
                                                country='Thailand'), 
                   exclude = c("s(country)","s(studyID)"), 
                   se.fit=T)
plot(x_parasites, 100*inv.logit(predspara_PCT$fit), ylab='mortality (%)', 
     xlab='', panel.first = grid(), 
     type='l', lwd=3,ylim = c(0,25), xaxt='n')
mtext(text = 'Parasitized RBCs (%)',side = 1,line = 2.5,cex=.7)
axis(1, at = c(-2,-1,0,log10(30)), labels = c(0.01,0.1,1,30))
axis(1, at = log10(c(10,20)), labels = NA)
polygon(c(x_parasites,rev(x_parasites)), 
        100*inv.logit(c(predspara_PCT$fit + 1.96*predspara_PCT$se.fit,
                        rev(predspara_PCT$fit - 1.96*predspara_PCT$se.fit))), 
        lty=2,lwd=2, border = NA, col = rgb(0, 0, 1,0.5))

#**************  Coma
predscoma = predict(modcoma, newdata = data.frame(coma=x_comas, 
                                                  studyID='AQ', 
                                                  country='Thailand'), 
                    exclude = c("s(country)","s(studyID)"), 
                    se.fit=T)
plot(x_comas, 100*inv.logit(predscoma$fit), ylab='mortality (%)', 
     panel.first = grid(), xlab='', 
     type='p', lwd=3, ylim = c(0,28), xaxt='n', 
     pch=19, xlim = c(-.25,1.25))
axis(1, at=x_comas, labels = c('No Coma','Coma'), 
     tick = F,line = .5,cex.axis=.9)
ys = 100*inv.logit(c(predscoma$fit - 1.96*predscoma$se.fit,
                     predscoma$fit + 1.96*predscoma$se.fit))
arrows(x0 = x_comas[1],y0 = ys[1], x1 = x_comas[1], y1 = ys[3],
       angle = 90,lwd=6,code = 3,length=.15,col = rgb(0, 0, 1,0.5))
arrows(x0 = x_comas[2],y0 = ys[2], x1 = x_comas[2], y1 = ys[4],
       angle = 90,lwd=6,code = 3,length=.15,col = rgb(0, 0, 1,0.5))


#************** Convulsions
predsconv = predict(modconv, newdata = data.frame(convulsions=x_conv, 
                                                  studyID='AQ', 
                                                  country='Thailand'), 
                    exclude = c("s(country)","s(studyID)"), 
                    se.fit=T)
plot(x_conv, 100*inv.logit(predsconv$fit), 
     panel.first = grid(), ylab='mortality (%)', xlab='', 
     type='p', lwd=3, ylim = c(0,30), xaxt='n',
     pch=19, xlim = c(-.25,1.25))
axis(1, at=x_conv, labels = c('No seizures','Seizures'),
     tick = F,line = .5,cex.axis=.9)

ys = 100*inv.logit(c(predsconv$fit + 1.96*predsconv$se.fit,
                     predsconv$fit -1.96*predsconv$se.fit))
arrows(x0 = x_conv[1],y0 = ys[1], x1 = x_conv[1], y1 = ys[3],angle = 90,
       lwd=6,code = 3,length=.15,col = rgb(0, 0, 1,0.5))
arrows(x0 = x_conv[2],y0 = ys[2], x1 = x_conv[2], y1 = ys[4],angle = 90,
       lwd=6,code = 3,length=.15,col = rgb(0, 0, 1,0.5))

#************** Pulmonary Oedema
predspoedema = predict(modpoedema, 
                       newdata = data.frame(poedema=x_poedema, 
                                            studyID='AQ', 
                                            country='Thailand'), 
                       exclude = c("s(country)","s(studyID)"), 
                       se.fit=T)
plot(x_poedema, 100*inv.logit(predspoedema$fit), 
     panel.first = grid(), ylab='mortality (%)', xlab='', 
     type='p', lwd=3, ylim = c(0,55), xaxt='n',
     pch=19, xlim = c(-.25,1.25))
axis(1, at=x_poedema, labels = c('No\nPulmonary\nOedema','Pulmonary\nOedema'),
     tick = F,line = 1.5,cex.axis=.9)
ys = 100*inv.logit(c(predspoedema$fit + 1.96*predspoedema$se.fit,
                     (predspoedema$fit - 1.96*predspoedema$se.fit)))
arrows(x0 = x_poedema[1],y0 = ys[1], x1 = x_poedema[1], y1 = ys[3],angle = 90,
       lwd=6,code = 3,length=.15,col = rgb(0, 0, 1,0.5))
arrows(x0 = x_poedema[2],y0 = ys[2], x1 = x_poedema[2], y1 = ys[4],angle = 90,
       lwd=6,code = 3,length=.15,col = rgb(0, 0, 1,0.5))

#************** Shock
predsshock = predict(modshock, 
                       newdata = data.frame(shock=x_shock, 
                                            studyID='AQ', 
                                            country='Thailand'), 
                       exclude = c("s(country)","s(studyID)"), 
                       se.fit=T)
plot(x_shock, 100*inv.logit(predsshock$fit), 
     panel.first = grid(), ylab='mortality (%)', xlab='', 
     type='p', lwd=3, ylim = c(0,27), xaxt='n',
     pch=19, xlim = c(-.25,1.25))
axis(1, at=x_shock, labels = c('No shock','Shock'),
     tick = F,line = 1.5,cex.axis=.9)
ys = 100*inv.logit(c(predsshock$fit + 1.96*predsshock$se.fit,
                     (predsshock$fit - 1.96*predsshock$se.fit)))
arrows(x0 = x_shock[1],y0 = ys[1], x1 = x_shock[1], y1 = ys[3],angle = 90,
       lwd=6,code = 3,length=.15,col = rgb(0, 0, 1,0.5))
arrows(x0 = x_shock[2],y0 = ys[2], x1 = x_shock[2], y1 = ys[4],angle = 90,
       lwd=6,code = 3,length=.15,col = rgb(0, 0, 1,0.5))
```

# Sensitivity Analysis

## Only using complete records

```{r}
vars_interest = c('outcome','HCT','LPAR_pct','coma','convulsions','poedema',
                  'BUN','BD','shock','hypoglycaemia',
                  'studyID','country')
complete_cases = apply(m[,vars_interest], 1, function(x) sum(is.na(x))) == 0
Complete_Leg_data = m[complete_cases,] 
Complete_Leg_data$studyID = as.factor(as.character(Complete_Leg_data$studyID))
```

```{r}
DAG_fmla = "outcome ~ HCT + LPAR_pct + coma + convulsions + poedema + 
              log2(BUN) + BD + shock + hypoglycaemia + 
              drug_AS + (1 | studyID) + (1 | country)"
# fit the model to each dataset
mod1 = glmer(formula = DAG_fmla, data = Complete_Leg_data, 
             family=binomial) 
summary(mod1)
```

```{r}
# The scalar multiples to put the AORs on the correct scales
# Compute 95% CIs
FixedEffs = as.data.frame(summary(mod1)$coefficients)
FixedEffs$Scalar_f = c(1,10, 1, 1, 1, 1,log2(3), 10, 1, 1,1)
Results = data.frame(lowerCI = exp(FixedEffs$Scalar_f*(FixedEffs$Estimate -
                                                         1.96*FixedEffs$`Std. Error`)),
                     mean = exp(FixedEffs$Scalar_f*(FixedEffs$Estimate)),
                     upperCI = exp(FixedEffs$Scalar_f*(FixedEffs$Estimate +
                                                         1.96*FixedEffs$`Std. Error`)))
rownames(Results) = rownames(FixedEffs)
```

Make the 'forest' plot:
```{r ForestPlot_CompleteData}
plotting_ind = rownames(Results) %in% c('BD','coma','convulsions1','LPAR_pct',
                                        'drug_AS','HCT','log2(BUN)','poedema1','shock1')
Results = Results[plotting_ind,]
Results$Y_Labels = c('-10 % points\nabsolute haematocrit\non admission',
                     'Tenfold increase\nin parasitaemia',
                     'Coma\non admission',
                     'Seizures\non admission',
                     'Pulmonary\nOedema\non admission',
                     '3 fold increase\nin blood urea\nnitrogen (mmol/L)',
                     '+10 mEq/L\nbase deficit',
                     'Shock\non admission',
                     'Artemisinin drug\nversus\nnon Artemisinin drug'
)

Results['HCT',1:3] = 1/Results['HCT',1:3]
x_ind = sort.int(Results$mean, index.return = T)$ix
Results = Results[x_ind,]
par(bty='n', las=1, mar = c(3,9,2,2))

xlims = c(0.5, 8.5)
plot(NA,NA, xlim= log2(xlims), ylim = c(0,1),xaxt='n',
     xlab='', ylab='', yaxt='n')
axis(1, at = log2(c(0.5,1, 2,4)), labels = c(0.5,1, 2,4))
abline(v=0, lty=2, lwd=3, col='red')
yindex =1
ypos = seq(0,1,length.out = sum(plotting_ind))


for(i in 1:nrow(Results)){
  arrows(log2(Results[i,'lowerCI']),ypos[yindex],
         log2(Results[i,'upperCI']),ypos[yindex],
         length=0.05, angle=90, code=3, 
         col = 'black',lwd=3)
  points(log2(Results[i,'mean']),ypos[yindex],pch=18,cex=2)
  yindex=yindex+1
  
}
abline(h=ypos, lty=3)
axis(side = 2, at = ypos, labels = Results$Y_Labels,tick=FALSE)
mtext(side=1, line = 2, text = 'Adjusted odds ratio')
mtext(side = 3, line = 1, text = 'Increased survival',adj = 0)
mtext(side = 3, line = 1, text = 'Decreased survival',adj = 1)
```



## The effect of Transfusion

We stratify by time to death.

Some data adjustments: need to do this elsewhere
Not open source data.

```{r, include=FALSE}
CMAL=readstata13::read.dta13('../../../Datasets/Malaria Core/DBallCore2016_V3.dta')
CMAL = CMAL[,c('StudyNumber','studyID','Timetodeathhrs')]
m$Unique_ID = apply(m, 1, function(x) paste(x['StudyNumber'], x['studyID'],sep='_'))
CMAL$Unique_ID = apply(CMAL, 1, function(x) paste(x['StudyNumber'], x['studyID'],sep='_'))

m = merge(m, CMAL[,c('Unique_ID','Timetodeathhrs')], by='Unique_ID')
time_death_NA = is.na(m$Timetodeathhrs) & m$outcome==1
m = m[!time_death_NA, ]
m$LPAR_pct[is.infinite(m$LPAR_pct)] = 0

library(haven)
seaqmat = read_sav("../../../Datasets/SEAQUAMAT/seaquamat stage 3.sav")
seaqmat$Unique_ID = apply(seaqmat, 1, function(x) paste(x['studyno'], 'SEAQUAMAT',sep='_'))
m = merge(m, seaqmat[,c('Unique_ID','transfusion')], by='Unique_ID', all = T)
m$transfusion = m$transfusion.x
m$transfusion[m$studyID=='SEAQUAMAT'] = m$transfusion.y[m$studyID=='SEAQUAMAT']


vars = c('HCT','LPAR_pct','coma' ,'convulsions',
         'poedema','BUN','BD' ,'shock','hypoglycaemia', 'transfusion')

```

Select the patients who survive past 4 hours
```{r}
#### We're just looking at the those who didn't die before 4 hours post admission
m$Outcome4hours = 0
m$Outcome4hours[!is.na(m$Timetodeathhrs) & m$Timetodeathhrs < 4] = 1

# We only look at individuals who survive past 4 hours
m = filter(m, Outcome4hours==0)
```


Multiple Imputation:
```{r}
###----- Do the imputation ---
# Base deficit from bicarbonate
BD_and_bicarbonate = !is.na(m$BD) & !is.na(m$bicarbonate)
mod_impute1 = lmer(BD ~ bicarbonate + (1 | studyID) + (1 | country), 
                   data= m[BD_and_bicarbonate,])
# Base deficit and lactate
BD_and_lactate = !is.na(m$BD) & !is.na(m$lactate)
mod_impute2 = lmer(BD ~ lactate + (1 | country), 
                   data= m[BD_and_lactate,])
# Base deficit and rr
BD_and_rr = !is.na(m$BD) & !is.na(m$rr)
mod_impute3 = lmer(BD ~ rr + (1 | studyID) + (1 | country), 
                   data= m[BD_and_rr,])

ind1 = is.na(m$BD) & !is.na(m$bicarbonate)
ind2 = !ind1 & is.na(m$BD) & !is.na(m$lactate)
ind3 = !ind1 & !ind2 & is.na(m$BD) & !is.na(m$rr)

# BUN and creatinine
BUN_and_cr = !is.na(m$BUN) & !is.na(m$creatinine)
mod_impute4 = lmer(log(BUN) ~ creatinine + (1 | studyID) + (1 | country), 
                   data= m[BUN_and_cr,])
ind4 = is.na(m$BUN) & !is.na(m$creatinine)

# Age from HCT
HCT_and_Age = !is.na(m$AgeInYear) & !is.na(m$HCT)
mod_impute5 = lmer(HCT ~ AgeInYear + (1 | studyID) + (1 | country),
                   data = m[HCT_and_Age,])
ind5 = is.na(m$HCT) & !is.na(m$AgeInYear)

# marginal distribution of BD to impute remaining
BD_alone = !is.na(m$BD)
mod_impute6 = lmer(BD ~ AgeInYear + (1 | studyID) + (1 | country),
                   data = m[BD_alone,])
ind6 = !ind1 & !ind2 & !ind3 & is.na(m$BD)

# marginal distribution of coma 
Coma_alone = !is.na(m$coma)
mod_impute7 = glmer(coma ~ AgeInYear + hypoglycaemia + (1 | studyID) + (1 | country),
                    family = 'binomial',
                    data = m[Coma_alone,])
ind7 = is.na(m$coma)

# marginal distribution of LPAR_pct 
LPAR_pct_alone = !is.na(m$LPAR_pct)
mod_impute8 = lmer(LPAR_pct ~ AgeInYear + (1 | studyID) + (1 | country),
                   data = m[LPAR_pct_alone,])
ind8 = is.na(m$LPAR_pct)

# marginal distribution of LPAR_pct 
Hypoglycaemia_alone = !is.na(m$hypoglycaemia)
mod_impute9 = glmer(hypoglycaemia ~ 1 + (1 | studyID) + (1 | country),
                    data = m[Hypoglycaemia_alone,], family='binomial')
ind9 = is.na(m$hypoglycaemia)

# marginal distribution of BUN 
BUN_alone = !is.na(m$BUN)
mod_impute10 = lmer(log(BUN) ~ 1 + (1 | studyID) + (1 | country),
                    data = m[BUN_alone,])
ind10 = !ind4 & is.na(m$BUN)

#transfusion from HCT
TF_and_HCT = !is.na(m$transfusion) & !is.na(m$HCT)
mod_impute0 = glmer(transfusion ~ HCT + (1 | studyID) + (1 | country), 
                    data= m[TF_and_HCT,], family = 'binomial')
ind0 = is.na(m$transfusion) & !is.na(m$HCT)

#----------------------------------------------------------------------
K_imputations = 100

SM_Impute_List = list()

for(k in 1:K_imputations){
  #print(k)
  Imp_data = m
  
  # BD from bicarbonate
  coefs1 = summary(mod_impute1)$coefficients
  Imp_data$BD[ind1] = rnorm(n = sum(ind1), 
                            mean = predict(mod_impute1, newdata = m[ind1,],
                                           allow.new.levels=T),
                            sd = coefs1[1,'Std. Error'] + 
                              coefs1[2,'Std. Error']*m$bicarbonate[ind1])
  
  # BD from lactate
  coefs2 = summary(mod_impute2)$coefficients
  Imp_data$BD[ind2] = rnorm(n = sum(ind2), 
                            mean = predict(mod_impute2, newdata = m[ind2,],allow.new.levels=T),
                            sd = coefs2[1,'Std. Error'] + 
                              coefs2[2,'Std. Error']*m$lactate[ind2])
  
  # BD from respiratory rate
  coefs3 = summary(mod_impute3)$coefficients
  Imp_data$BD[ind3] = rnorm(n = sum(ind3), 
                            mean = predict(mod_impute3, newdata = m[ind3,],allow.new.levels=T),
                            sd = coefs3[1,'Std. Error'] + 
                              coefs3[2,'Std. Error']*m$rr[ind3])
  
  # BUN from creatinine
  coefs4 = summary(mod_impute4)$coefficients
  Imp_data$BUN[ind4] = exp(rnorm(n = sum(ind4), 
                                 mean = predict(mod_impute4, newdata = m[ind4,],allow.new.levels=T),
                                 sd = coefs4[1,'Std. Error'] + 
                                   coefs4[2,'Std. Error']*m$creatinine[ind4]))
  
  # HCT from Age
  coefs5 = summary(mod_impute5)$coefficients
  Imp_data$HCT[ind5] = rnorm(n = sum(ind5), 
                             mean = predict(mod_impute5, newdata = m[ind5,],
                                            allow.new.levels=T),
                             sd = coefs5[1,'Std. Error'] + 
                               coefs5[2,'Std. Error']*m$AgeInYear[ind5])
  
  # BD from age
  coefs6 = summary(mod_impute6)$coefficients
  Imp_data$BD[ind6] = rnorm(n = sum(ind6), 
                            mean = predict(mod_impute6, newdata = m[ind6,],
                                           allow.new.levels=T),
                            sd = coefs6[1,'Std. Error'] + 
                              coefs6[2,'Std. Error']*m$AgeInYear[ind6])
  
  # Coma from hypoglycaemia
  coefs7 = summary(mod_impute7)$coefficients
  Imp_data$coma[ind7] = rbinom(n = sum(ind7), size = 1,
                               predict(mod_impute7, newdata = m[ind7,],
                                       allow.new.levels=T, type='response'))
  
  # Parasitaemia from age
  coefs8 = summary(mod_impute8)$coefficients
  Imp_data$LPAR_pct[ind8] = rnorm(n = sum(ind8), 
                                  mean = predict(mod_impute8, 
                                                 newdata = m[ind8,],
                                                 allow.new.levels=T),
                                  sd = coefs8[1,'Std. Error'] + 
                                    coefs8[2,'Std. Error']*m$AgeInYear[ind8])
  
  # Hypoglycaemia: marginal
  coefs9 = summary(mod_impute9)$coefficients
  Imp_data$hypoglycaemia[ind9] = rbinom(n = sum(ind9),size = 1, 
                                        prob = predict(mod_impute9, 
                                                       newdata = m[ind9,],
                                                       allow.new.levels=T,
                                                       type='response'))
  
  # BUN: marginal
  coefs10 = summary(mod_impute10)$coefficients
  Imp_data$BUN[ind10] = exp(rnorm(n = sum(ind10), 
                                  mean = predict(mod_impute10, newdata = m[ind10,],
                                                 allow.new.levels=T),
                                  sd = coefs10[1,'Std. Error']))
  
  # Transfusion from HCT
  coefs0 = summary(mod_impute0)$coefficients
  Imp_data$HCT[ind0] = rbinom(n = sum(ind0), size = 1,
                              predict(mod_impute0, newdata = m[ind0,],
                                      allow.new.levels=T, type='response'))
  
  SM_Impute_List[[k]] = Imp_data
  
}


```

```{r}
if(RUN_MODELS){
  DAG_fmla = "outcome ~ HCT + LPAR_pct + coma + convulsions + poedema + 
                      log2(BUN) + BD + shock + hypoglycaemia + 
                      transfusion +
                      drug_AS + (1 | studyID) + (1 | country)"
  # fit the model to each dataset
  modList= glmerModList(DAG_fmla,data = SM_Impute_List, 
                        family=binomial,parallel = T) 
  save(modList, file = 'RData/ModelList_Transfusion.RData')
} else{
  load(file = 'RData/ModelList_Transfusion.RData')
}

FixedEffs = modelFixedEff(modList)
RandEffs = modelRandEffStats(modList)
print(FixedEffs)
print(RandEffs)
```



```{r ForestPlot_TransfusionAdjusted, echo=FALSE}
# The scalar multiples to put the AORs on the correct scales
Scalar_f = c(1, 10, 1, 1, 1, 10, 1, log2(3), 1, 1, 1, 1)
# Compute 95% CIs
Results = data.frame(lowerCI = exp(Scalar_f*(FixedEffs$estimate -
                                               1.96*FixedEffs$std.error)),
                     mean = exp(Scalar_f*(FixedEffs$estimate)),
                     upperCI = exp(Scalar_f*(FixedEffs$estimate +
                                               1.96*FixedEffs$std.error)))
rownames(Results) = FixedEffs$term
Results['HCT',] = 1/Results['HCT',]
Results

Results = Results[!(rownames(Results) %in% c('hypoglycaemia','(Intercept)')),]
rownames(Results) = c('+10 mEq/L\nbase deficit',
                      'Coma\non admission',
                      'Seizures\non admission',
                      'Artemisinin drug\nversus\nnon Artemisinin drug',
                      '-10 % points\nabsolute haematocrit\non admission',
                      '3 fold increase\nin blood urea\nnitrogen (mmol/L)',
                      'Tenfold increase\nin parasitaemia',
                      'Pulmonary\nOedema\non admission',
                      'Shock\non admission',
                      'Transfusion')


x_ind = sort.int(Results$mean, index.return = T)$ix
Results = Results[x_ind,]
par(bty='n', las=1, mar = c(4,9,2,2))

xlims = c(0.5, 8)
plot(NA,NA, xlim= log2(xlims), ylim = c(0,1),xaxt='n',
     xlab='', ylab='', yaxt='n')
axis(1, at = log2(c(0.5,1, 2,4,8)), labels = c(0.5,1, 2,4,8))
abline(v=0, lty=2, lwd=3, col='red')
yindex =1
ypos = seq(0,1,length.out = nrow(Results))


for(i in 1:nrow(Results)){
  arrows(log2(Results[i,'lowerCI']),ypos[yindex],
         log2(Results[i,'upperCI']),ypos[yindex],
         length=0.05, angle=90, code=3, 
         col = 'black',lwd=3)
  points(log2(Results[i,'mean']),ypos[yindex],pch=18,cex=2)
  yindex=yindex+1
  
}
abline(h=ypos, lty=3)
axis(side = 2, at = ypos, labels = rownames(Results),tick=FALSE)
mtext(side=1, line = 2.5, text = 'Adjusted odds ratio')
mtext(side = 3, line = 1, text = 'Increased survival',adj = 0)
mtext(side = 3, line = 1, text = 'Decreased survival',adj = 1)

```


## Effect of unmeasured sepsis

Select the patients in Asia (no concomitant sepsis):

```{r}
load('RData/Data.RData')
m$coma[is.na(m$coma) & m$studyID=='AQ'] = 0

m$drug_AS = 0
m$drug_AS[m$drug_class=='artemisinin']=1
m$LPAR_pct[is.infinite(m$LPAR_pct)] = 0

m = filter(m, country %in% c('Bangladesh','India','Indonesia','Myanmar','Thailand','Vietnam'))
```


Multiple Imputation:
```{r}
###----- Do the imputation and get the indices for imputation
# Base deficit and lactate
BD_and_lactate = !is.na(m$BD) & !is.na(m$lactate)
mod_impute2 = lmer(BD ~ lactate + (1 | country), 
                   data= m[BD_and_lactate,])
# Base deficit and rr
BD_and_rr = !is.na(m$BD) & !is.na(m$rr)
mod_impute3 = lmer(BD ~ rr + (1 | studyID) + (1 | country), 
                   data= m[BD_and_rr,])

ind1 = is.na(m$BD) & !is.na(m$bicarbonate)
ind2 = !ind1 & is.na(m$BD) & !is.na(m$lactate)
ind3 = !ind1 & !ind2 & is.na(m$BD) & !is.na(m$rr)

# BUN and creatinine
BUN_and_cr = !is.na(m$BUN) & !is.na(m$creatinine)
mod_impute4 = lmer(log(BUN) ~ creatinine + (1 | studyID) + (1 | country), 
                   data= m[BUN_and_cr,])
ind4 = is.na(m$BUN) & !is.na(m$creatinine)

# Age from HCT
HCT_and_Age = !is.na(m$AgeInYear) & !is.na(m$HCT)
mod_impute5 = lmer(HCT ~ AgeInYear + (1 | studyID) + (1 | country),
                   data = m[HCT_and_Age,])
ind5 = is.na(m$HCT) & !is.na(m$AgeInYear)

# marginal distribution of BD to impute remaining
BD_alone = !is.na(m$BD)
mod_impute6 = lmer(BD ~ AgeInYear + (1 | studyID) + (1 | country),
                   data = m[BD_alone,])
ind6 = !ind1 & !ind2 & !ind3 & is.na(m$BD)

# marginal distribution of coma 
Coma_alone = !is.na(m$coma)
mod_impute7 = glmer(coma ~ AgeInYear + hypoglycaemia + (1 | studyID) + (1 | country),
                    family = 'binomial',
                    data = m[Coma_alone,])
ind7 = is.na(m$coma)

# marginal distribution of LPAR_pct 
LPAR_pct_alone = !is.na(m$LPAR_pct)
mod_impute8 = lmer(LPAR_pct ~ AgeInYear + (1 | studyID) + (1 | country),
                   data = m[LPAR_pct_alone,])
ind8 = is.na(m$LPAR_pct)

# marginal distribution of LPAR_pct 
Hypoglycaemia_alone = !is.na(m$hypoglycaemia)
mod_impute9 = glmer(hypoglycaemia ~ 1 + (1 | studyID) + (1 | country),
                    data = m[Hypoglycaemia_alone,], family='binomial')
ind9 = is.na(m$hypoglycaemia)

# marginal distribution of BUN 
BUN_alone = !is.na(m$BUN)
mod_impute10 = lmer(log(BUN) ~ 1 + (1 | studyID) + (1 | country),
                    data = m[BUN_alone,])
ind10 = !ind4 & is.na(m$BUN)

#transfusion from HCT
TF_and_HCT = !is.na(m$transfusion) & !is.na(m$HCT)
mod_impute0 = glmer(transfusion ~ HCT + (1 | studyID) + (1 | country), 
                    data= m[TF_and_HCT,], family = 'binomial')
ind0 = is.na(m$transfusion) & !is.na(m$HCT)

#----------------------------------------------------------------------
K_imputations = 100

SM_Impute_List = list()

for(k in 1:K_imputations){
  #print(k)
  Imp_data = m
  
  # BD from lactate
  coefs2 = summary(mod_impute2)$coefficients
  Imp_data$BD[ind2] = rnorm(n = sum(ind2), 
                            mean = predict(mod_impute2, newdata = m[ind2,],allow.new.levels=T),
                            sd = coefs2[1,'Std. Error'] + 
                              coefs2[2,'Std. Error']*m$lactate[ind2])
  
  # BD from respiratory rate
  coefs3 = summary(mod_impute3)$coefficients
  Imp_data$BD[ind3] = rnorm(n = sum(ind3), 
                            mean = predict(mod_impute3, newdata = m[ind3,],allow.new.levels=T),
                            sd = coefs3[1,'Std. Error'] + 
                              coefs3[2,'Std. Error']*m$rr[ind3])
  
  # BUN from creatinine
  coefs4 = summary(mod_impute4)$coefficients
  Imp_data$BUN[ind4] = exp(rnorm(n = sum(ind4), 
                                 mean = predict(mod_impute4, newdata = m[ind4,],allow.new.levels=T),
                                 sd = coefs4[1,'Std. Error'] + 
                                   coefs4[2,'Std. Error']*m$creatinine[ind4]))
  
  # HCT from Age
  coefs5 = summary(mod_impute5)$coefficients
  Imp_data$HCT[ind5] = rnorm(n = sum(ind5), 
                             mean = predict(mod_impute5, newdata = m[ind5,],
                                            allow.new.levels=T),
                             sd = coefs5[1,'Std. Error'] + 
                               coefs5[2,'Std. Error']*m$AgeInYear[ind5])
  
  # BD from age
  coefs6 = summary(mod_impute6)$coefficients
  Imp_data$BD[ind6] = rnorm(n = sum(ind6), 
                            mean = predict(mod_impute6, newdata = m[ind6,],
                                           allow.new.levels=T),
                            sd = coefs6[1,'Std. Error'] + 
                              coefs6[2,'Std. Error']*m$AgeInYear[ind6])
  
  # Coma from hypoglycaemia
  coefs7 = summary(mod_impute7)$coefficients
  Imp_data$coma[ind7] = rbinom(n = sum(ind7), size = 1,
                               predict(mod_impute7, newdata = m[ind7,],
                                       allow.new.levels=T, type='response'))
  
  # Parasitaemia from age
  coefs8 = summary(mod_impute8)$coefficients
  Imp_data$LPAR_pct[ind8] = rnorm(n = sum(ind8), 
                                  mean = predict(mod_impute8, 
                                                 newdata = m[ind8,],
                                                 allow.new.levels=T),
                                  sd = coefs8[1,'Std. Error'] + 
                                    coefs8[2,'Std. Error']*m$AgeInYear[ind8])
  
  # Hypoglycaemia: marginal
  coefs9 = summary(mod_impute9)$coefficients
  Imp_data$hypoglycaemia[ind9] = rbinom(n = sum(ind9),size = 1, 
                                        prob = predict(mod_impute9, 
                                                       newdata = m[ind9,],
                                                       allow.new.levels=T,
                                                       type='response'))
  
  # BUN: marginal
  coefs10 = summary(mod_impute10)$coefficients
  Imp_data$BUN[ind10] = exp(rnorm(n = sum(ind10), 
                                  mean = predict(mod_impute10, newdata = m[ind10,],
                                                 allow.new.levels=T),
                                  sd = coefs10[1,'Std. Error']))
  
  # Transfusion from HCT
  coefs0 = summary(mod_impute0)$coefficients
  Imp_data$HCT[ind0] = rbinom(n = sum(ind0), size = 1,
                              predict(mod_impute0, newdata = m[ind0,],
                                      allow.new.levels=T, type='response'))
  
  SM_Impute_List[[k]] = Imp_data
  
}


```

run the models on the list of imputed datsets
```{r}
if(RUN_MODELS){
  DAG_fmla = "outcome ~ HCT + LPAR_pct + coma + convulsions + poedema + 
                      log2(BUN) + BD + shock + hypoglycaemia + 
                      drug_AS + (1 | studyID) + (1 | country)"
  # fit the model to each dataset
  modList= glmerModList(DAG_fmla,data = SM_Impute_List, 
                        family=binomial,parallel = T) 
  save(modList, file = 'RData/ModelList_Asia.RData')
} else{
  load(file = 'RData/ModelList_Asia.RData')
}

FixedEffs = modelFixedEff(modList)
RandEffs = modelRandEffStats(modList)
print(FixedEffs)
print(RandEffs)
```



```{r ForestPlot_Asia, echo=FALSE}
# The scalar multiples to put the AORs on the correct scales
Scalar_f = c(1, 10, 1, 1, 1, 10, 1, log2(3), 1, 1, 1)
# Compute 95% CIs
Results = data.frame(lowerCI = exp(Scalar_f*(FixedEffs$estimate -
                                               1.96*FixedEffs$std.error)),
                     mean = exp(Scalar_f*(FixedEffs$estimate)),
                     upperCI = exp(Scalar_f*(FixedEffs$estimate +
                                               1.96*FixedEffs$std.error)))
rownames(Results) = FixedEffs$term
Results




Results['HCT',] = 1/Results['HCT',]
Results = Results[!(rownames(Results) %in% c('hypoglycaemia','(Intercept)')),]
rownames(Results) = c('+10 mEq/L\nbase deficit',
                      'Coma\non admission',
                      'Seizures\non admission',
                      'Artemisinin drug\nversus\nnon Artemisinin drug',
                      '-10 % points\nabsolute haematocrit\non admission',
                      '3 fold increase\nin blood urea\nnitrogen (mmol/L)',
                      'Tenfold increase\nin parasitaemia',
                      'Pulmonary\nOedema\non admission',
                      'Shock\non admission')

x_ind = sort.int(Results$mean, index.return = T)$ix
Results = Results[x_ind,]
par(bty='n', las=1, mar = c(4,9,2,2))


xlims = c(0.5,4)
plot(NA,NA, xlim= log2(xlims), ylim = c(0,1),xaxt='n',
     xlab='', ylab='', yaxt='n')
axis(1, at = log2(c(0.5,1, 2,4)), labels = c(0.5,1, 2,4))
abline(v=0, lty=2, lwd=3, col='red')
yindex =1
ypos = seq(0,1,length.out = nrow(Results))


for(i in 1:nrow(Results)){
  arrows(log2(Results[i,'lowerCI']),ypos[yindex],
         log2(Results[i,'upperCI']),ypos[yindex],
         length=0.05, angle=90, code=3, 
         col = 'black',lwd=3)
  points(log2(Results[i,'mean']),ypos[yindex],pch=18,cex=2)
  yindex=yindex+1
  
}
abline(h=ypos, lty=3)
axis(side = 2, at = ypos, labels = rownames(Results),tick=FALSE)
mtext(side=1, line = 2.5, text = 'Adjusted odds ratio')
mtext(side = 3, line = 1, text = 'Increased survival',adj = 0)
mtext(side = 3, line = 1, text = 'Decreased survival',adj = 1)

```

