---
title: "Charactersing effect of anaemia on mortality in severe malaria"
output:
  html_document:
    toc: true
    fig_caption: yes
    keep_md: yes
---

```{r, echo=F}
knitr::opts_chunk$set(cache = TRUE, cache.comments = FALSE, 
                      echo = TRUE, include = TRUE, 
                      fig.width = 7, fig.height = 7,
                      fig.pos = 'H', 
                      dev = 'cairo_pdf', dpi = 400)

RUN_MODELS = F
```

# Background

This looks at the severe malaria legacy dataset from MORU

```{r, echo=TRUE}
load('RData/Data.RData')
m$drug_AS = 0
m$drug_AS[m$drug_class=='artemisinin']=1
m$country = as.factor(as.character(m$country))
m$poedema[m$continent=='Africa']=NA
```




```{r, echo=TRUE, include=FALSE}
library(lme4)
require(missForest)
library(merTools)
library(fields)
library(plyr)
library(dplyr)
```



Data summaries
```{r}
writeLines(paste('Children less than 12 years:',
                 sum(m$AgeInYear < 12)))
writeLines(paste('Adults:',
                 sum(m$AgeInYear >= 12)))


writeLines(paste('Children in Africa:',
                 sum(m$AgeInYear < 12 & m$continent == 'Africa')))
writeLines(paste('Adults in Africa:',
                 sum(m$AgeInYear >= 12 & m$continent== 'Africa')))

writeLines(paste('Children in Asia:',
                 sum(m$AgeInYear < 12 & m$continent == 'Asia')))
writeLines(paste('Adults in Asia:',
                 sum(m$AgeInYear >= 12 & m$continent == 'Asia')))
writeLines('\nNumber of patients per study:\n')
table(m$studyID)

writeLines('\nStudies broken down by country:\n')

table(m$country,m$studyID)


for(ss in unique(m$studyID)){
  xx = filter(m, studyID==ss)
  writeLines(sprintf('\nThe mortality in %s was %s%%', ss, round(mean(xx$outcome)*100)))
}
```

# Multiple imputation using linear relationships

The number of missing variables in the pooled data:
```{r}
round(apply(m,2, function(x) sum(is.na(x))),0)

round(apply(m,2, function(x) 100*sum(is.na(x)))/nrow(m),0)
```

We make a few data adjustments for the model imputation and fitting:
```{r}
m$LPAR_pct[is.infinite(m$LPAR_pct)] = 0
```

We run the multiple imputation using chained equations.

Imputation models V2
```{r}

```

First we setup these linear models. Random effect terms are added for study and country.

```{r, include=FALSE}
# Pulmonary Oedema from rr
Poedema_rr = !is.na(m$poedema) & !is.na(m$rr)
mod_impute0 = glmer(poedema ~ rr + (1 | studyID), 
                   data= m[Poedema_rr,], family = 'binomial')
ind0 = is.na(m$poedema) & !is.na(m$rr)

# Pulmonary Oedema : marginal
Poedema_rr2 = !is.na(m$poedema)
mod_impute0b = glmer(poedema ~ (1 | studyID), 
                   data= m[Poedema_rr2,], family = 'binomial')
ind0b = is.na(m$poedema) & is.na(m$rr)

# Base deficit from bicarbonate
BD_and_bicarbonate = !is.na(m$BD) & !is.na(m$bicarbonate)
mod_impute1 = lmer(BD ~ bicarbonate + (1 | studyID) , 
                   data= m[BD_and_bicarbonate,])
# Base deficit and lactate
BD_and_lactate = !is.na(m$BD) & !is.na(m$lactate)
mod_impute2 = lm(BD ~ lactate, 
                   data= m[BD_and_lactate,])
# Base deficit and rr
BD_and_rr = !is.na(m$BD) & !is.na(m$rr)
mod_impute3 = lmer(BD ~ rr + (1 | studyID), 
                   data= m[BD_and_rr,])

ind1 = is.na(m$BD) & !is.na(m$bicarbonate)
ind2 = !ind1 & is.na(m$BD) & !is.na(m$lactate)
ind3 = !ind1 & !ind2 & is.na(m$BD) & !is.na(m$rr)

# BUN and creatinine
BUN_and_cr = !is.na(m$BUN) & !is.na(m$creatinine)
mod_impute4 = lmer(log(BUN) ~ creatinine + (1 | studyID) + (1 | country), 
                   data= m[BUN_and_cr,])
ind4 = is.na(m$BUN) & !is.na(m$creatinine)

# Age from HCT
HCT_and_Age = !is.na(m$AgeInYear) & !is.na(m$HCT)
mod_impute5 = lmer(HCT ~ AgeInYear + (1 | studyID) + (1 | country),
                   data = m[HCT_and_Age,])
ind5 = is.na(m$HCT) & !is.na(m$AgeInYear)

# marginal distribution of BD to impute remaining
BD_alone = !is.na(m$BD)
mod_impute6 = lmer(BD ~ AgeInYear + (1 | studyID) + (1 | country),
                   data = m[BD_alone,])
ind6 = !ind1 & !ind2 & !ind3 & is.na(m$BD)

# marginal distribution of coma 
Coma_alone = !is.na(m$coma)
mod_impute7 = glmer(coma ~ AgeInYear + hypoglycaemia + (1 | studyID),
                    family = 'binomial',
                    data = m[Coma_alone,])
ind7 = is.na(m$coma) & !is.na(m$hypoglycaemia)

# marginal distribution of coma 
mod_impute7b = glmer(coma ~ AgeInYear + (1 | studyID),
                     family = 'binomial',
                     data = m[Coma_alone,])
ind7b = is.na(m$coma) & is.na(m$hypoglycaemia)

# marginal distribution of LPAR_pct 
LPAR_pct_alone = !is.na(m$LPAR_pct)
mod_impute8 = lmer(LPAR_pct ~ AgeInYear + (1 | studyID) + (1 | country),
                   data = m[LPAR_pct_alone,])
ind8 = is.na(m$LPAR_pct)

# marginal distribution of LPAR_pct 
Hypoglycaemia_alone = !is.na(m$hypoglycaemia)
mod_impute9 = glmer(hypoglycaemia ~ 1 + (1 | studyID) + (1 | country),
                    data = m[Hypoglycaemia_alone,], family='binomial')
ind9 = is.na(m$hypoglycaemia)

# marginal distribution of BUN 
BUN_alone = !is.na(m$BUN)
mod_impute10 = lmer(log(BUN) ~ 1 + (1 | studyID) + (1 | country),
                    data = m[BUN_alone,])
ind10 = !ind4 & is.na(m$BUN)
```


```{r, include=FALSE}
if(RUN_MODELS){
  K_imputations = 100
  SM_Impute_List = list()
  for (k in 1:K_imputations){
    Imp_data = m
    # Poedema from rr
    coefs1 = summary(mod_impute0)$coefficients
    Imp_data$poedema[ind0] = rbinom(n = sum(ind0), size = 1,
                                  predict(mod_impute0, newdata = m[ind0,],
                                          allow.new.levels=T, type='response'))
    # Poedema marginal
    coefs1 = summary(mod_impute0b)$coefficients
    Imp_data$poedema[ind0b] = rbinom(n = sum(ind0b), size = 1,
                                  predict(mod_impute0b, newdata = m[ind0b,],
                                          allow.new.levels=T, type='response'))
    
    # BD from bicarbonate
    coefs1 = summary(mod_impute1)$coefficients
    Imp_data$BD[ind1] = rnorm(n = sum(ind1), 
                              mean = predict(mod_impute1, newdata = m[ind1,]),
                              sd = coefs1[1,'Std. Error'] + 
                                coefs1[2,'Std. Error']*m$bicarbonate[ind1])
    
    # BD from lactate
    coefs2 = summary(mod_impute2)$coefficients
    Imp_data$BD[ind2] = rnorm(n = sum(ind2), 
                              mean = predict(mod_impute2, 
                                             newdata = m[ind2,],
                                             allow.new.levels=T),
                              sd = coefs2[1,'Std. Error'] + 
                                coefs2[2,'Std. Error']*m$lactate[ind2])
    
    # BD from respiratory rate
    coefs3 = summary(mod_impute3)$coefficients
    Imp_data$BD[ind3] = rnorm(n = sum(ind3), 
                              mean = predict(mod_impute3, 
                                             newdata = m[ind3,],allow.new.levels=T),
                              sd = coefs3[1,'Std. Error'] + 
                                coefs3[2,'Std. Error']*m$rr[ind3])
    
    # BUN from creatinine
    coefs4 = summary(mod_impute4)$coefficients
    Imp_data$BUN[ind4] = exp(rnorm(n = sum(ind4), 
                                   mean = predict(mod_impute4, 
                                                  newdata = m[ind4,],
                                                  allow.new.levels=T),
                                   sd = coefs4[1,'Std. Error'] + 
                                     coefs4[2,'Std. Error']*m$creatinine[ind4]))
    
    # HCT from Age
    coefs5 = summary(mod_impute5)$coefficients
    Imp_data$HCT[ind5] = rnorm(n = sum(ind5), 
                               mean = predict(mod_impute5, 
                                              newdata = m[ind5,],
                                              allow.new.levels=T),
                               sd = coefs5[1,'Std. Error'] + 
                                 coefs5[2,'Std. Error']*m$AgeInYear[ind5])
    
    # BD from age
    coefs6 = summary(mod_impute6)$coefficients
    Imp_data$BD[ind6] = rnorm(n = sum(ind6), 
                              mean = predict(mod_impute6, 
                                             newdata = m[ind6,],
                                             allow.new.levels=T),
                              sd = coefs6[1,'Std. Error'] + 
                                coefs6[2,'Std. Error']*m$AgeInYear[ind6])
    
    # Coma from hypoglycaemia
    coefs7 = summary(mod_impute7)$coefficients
    Imp_data$coma[ind7] = rbinom(n = sum(ind7), size = 1,
                                 predict(mod_impute7, newdata = m[ind7,],
                                         allow.new.levels=T, type='response'))
    
    # Coma from age
    coefs7b = summary(mod_impute7b)$coefficients
    Imp_data$coma[ind7b] = rbinom(n = sum(ind7b), size = 1,
                                  predict(mod_impute7b, newdata = m[ind7b,],
                                          allow.new.levels=T, type='response'))
    
    # Parasitaemia from age
    coefs8 = summary(mod_impute8)$coefficients
    Imp_data$LPAR_pct[ind8] = rnorm(n = sum(ind8), 
                                    mean = predict(mod_impute8, 
                                                   newdata = m[ind8,],
                                                   allow.new.levels=T),
                                    sd = coefs8[1,'Std. Error'] + 
                                      coefs8[2,'Std. Error']*m$AgeInYear[ind8])
    
    # Hypoglycaemia: marginal
    coefs9 = summary(mod_impute9)$coefficients
    Imp_data$hypoglycaemia[ind9] = rbinom(n = sum(ind9),size = 1, 
                                          prob = predict(mod_impute9, 
                                                         newdata = m[ind9,],
                                                         allow.new.levels=T,
                                                         type='response'))
    
    # BUN: marginal
    coefs10 = summary(mod_impute10)$coefficients
    Imp_data$BUN[ind10] = exp(rnorm(n = sum(ind10), 
                                    mean = predict(mod_impute10, newdata = m[ind10,],
                                                   allow.new.levels=T),
                                    sd = coefs10[1,'Std. Error']))
    
    SM_Impute_List[[k]] = Imp_data
    print(k)
  }
  save(SM_Impute_List, file = 'RData/Multiple_Imputed_Datasets.RData')
  # Check the number of remaining missing in the last imputed dataset:
  vars_explanatory = c('HCT','LPAR_pct','coma' ,'convulsions',
                       'poedema','BUN','BD' ,'shock','hypoglycaemia',
                       'drug_AS','studyID','country')
  apply(Imp_data[,vars_explanatory], 2, function(x) sum(is.na(x)))
} else {
  load('RData/Multiple_Imputed_Datasets.RData')
}
```


# Logistic regression model

We fit the full model with adjustments as specified in the Methods section:
```{r deathmodel_glm, include=FALSE, echo=TRUE}
if(RUN_MODELS){
  # the model formula
  DAG_fmla = "outcome ~ HCT + LPAR_pct + coma + convulsions + poedema + 
                        log2(BUN) + BD + shock + hypoglycaemia + 
                        drug_AS + (1 | continent/country)"
  # fit the model to each dataset
  modList= glmerModList(DAG_fmla, data = SM_Impute_List,
                        family=binomial,parallel = T) 
  # save the output
  save(modList, file = 'RData/Models_List.RData')
} else {
  load('RData/Models_List.RData')
}
```

Compute the overall parameter estimates:
```{r}
# extract the fixed and random effects from all the model fits
# These functions then compute the overall estimates
FixedEffs = modelFixedEff(modList)
RandEffs = modelRandEffStats(modList)
print(FixedEffs)
print(RandEffs)


FixedEffs$Scalar_f = as.numeric(mapvalues(x = FixedEffs$term, 
                                          from = c("(Intercept)",
                                                   "BD",
                                                   "coma",
                                                   "convulsions1",
                                                   "drug_AS",
                                                   "HCT",
                                                   "hypoglycaemia",
                                                   "log2(BUN)",
                                                   "LPAR_pct",
                                                   "shock1",
                                                   "poedema1"),
                                          to = as.numeric(c(1, 7, 1, 1, 1, 10, 1,2,log10(6),1, 1))))
FixedEffs$std.error=as.numeric(FixedEffs$std.error)
FixedEffs$estimate=as.numeric(FixedEffs$estimate)

# Compute 95% CIs
Results = data.frame(lowerCI = exp(FixedEffs$Scalar_f*(FixedEffs$estimate -
                                                         1.96*FixedEffs$std.error)),
                     mean = exp(FixedEffs$Scalar_f*(FixedEffs$estimate)),
                     upperCI = exp(FixedEffs$Scalar_f*(FixedEffs$estimate +
                                                         1.96*FixedEffs$std.error)))
rownames(Results) = FixedEffs$term
```

Aggregate results for plotting:
```{r}
writeLines(sprintf('Standard deviation of base deficit in patient population is %s',
                   round(sd(m$BD,na.rm=T))))
writeLines(sprintf('Standard deviation of blood urea nitrogen in patient population is %s',
                   round(2^sd(log2(m$BUN),na.rm=T))))
writeLines(sprintf('Standard deviation of parasitised RBCs in patient population is %s',
                   round(10^sd(m$LPAR_pct,na.rm=T))))
writeLines(sprintf('Standard deviation of haematocrits in patient population is %s',
                   round(sd(m$HCT,na.rm=T))))
```



Make the 'forest' plot:
```{r ForestPlot_SM}
Results = Results[rownames(Results) %in% c('BD','coma','convulsions1','drug_AS','HCT',
                                           'log2(BUN)','poedema1','LPAR_pct','shock1'),]
Results$Names =mapvalues(rownames(Results),
                         from = c("drug_AS",
                                  "HCT",
                                  "LPAR_pct",
                                  "convulsions1",
                                  "shock1",
                                  "poedema1",
                                  "log2(BUN)",
                                  "BD", 
                                  "coma"),
                         to = c('Artemisinin drug\nversus\nnon Artemisinin drug',
                                '-10 % points\nabsolute haematocrit\non admission',
                                'Six fold increase\n in parasitised\nred blood cells',
                                'Seizures\non admission',
                                'Shock\non admission',
                                'Pulmonary\noedema\non admission',
                                'Two fold increase\nin blood urea\nnitrogen (mmol/L)',
                                '+7 mEq/L\nbase deficit',
                                'Coma\non admission'))
Results['HCT',c("lowerCI","mean","upperCI")] = 1/Results['HCT',c("lowerCI","mean","upperCI")]
Results = arrange(Results, mean)
print(round(Results[,c("lowerCI","mean","upperCI")],2))

par(bty='n', las=1, mar = c(4,9,2,2))

xlims = c(0.5, 4.5)
plot(NA,NA, xlim= log2(xlims), ylim = c(0,1),xaxt='n',
     xlab='', ylab='', yaxt='n')
abline(v= seq(-1,3,by=1),col = "lightgray", lty = "dotted",lwd = par("lwd"))
axis(1, at = log2(c(0.5,1, 2,4)), labels = c(0.5,1, 2,4))
abline(v=0, lty=2, lwd=3, col='red')
yindex =1
ypos = seq(0,1,length.out = nrow(Results))


for(i in 1:nrow(Results)){
  arrows(log2(Results[i,'lowerCI']),ypos[yindex],
         log2(Results[i,'upperCI']),ypos[yindex],
         length=0.0, angle=90, code=3, 
         col = 'black',lwd=3)
  points(log2(Results[i,'mean']),ypos[yindex],pch=18,cex=2)
  yindex=yindex+1
  
}
abline(h=ypos, lty=3)
axis(side = 2, at = ypos, labels = Results$Names,tick=FALSE)
mtext(side=1, line = 2.5, text = 'Adjusted odds ratio')
mtext(side = 3, line = 1, text = 'Increased survival',adj = 0)
mtext(side = 3, line = 1, text = 'Decreased survival',adj = 1)
```


# Logistic regression model with subsets of the data

### Only Asian patients

This provides a sensitivity analysis regarding confounding by sepsis.

```{r}
SM_Impute_List_Asia = list()
for (i in 1:length(SM_Impute_List)){
  SM_Impute_List_Asia[[i]] = filter(SM_Impute_List[[i]], continent == 'Asia')
}
```

We fit the full model with adjustments as specified in the Methods section:
```{r deathmodel_glm_Asia, include=FALSE, echo=TRUE}
if(RUN_MODELS){
  # the model formula
  DAG_fmla = "outcome ~ HCT + LPAR_pct + coma + convulsions + poedema + 
                      log2(BUN) + BD + shock + hypoglycaemia + 
                      drug_AS + (1 | country)"
  # fit the model to each dataset
  modList = glmerModList(DAG_fmla, data = SM_Impute_List_Asia,
                         family=binomial,parallel = T) 
  # save the output
  save(modList, file = 'RData/ModelList_Asia.RData')
} else {
  load('RData/ModelList_Asia.RData')
}

FixedEffs_Asia = modelFixedEff(modList)
print(FixedEffs_Asia)

FixedEffs_Asia$Scalar_f = as.numeric(mapvalues(x = FixedEffs_Asia$term, 
                                               from = c("(Intercept)",
                                                        "BD",
                                                        "coma",
                                                        "convulsions1",
                                                        "drug_AS",
                                                        "HCT",
                                                        "hypoglycaemia",
                                                        "log2(BUN)",
                                                        "LPAR_pct",
                                                        "shock1",
                                                        "poedema1"),
                                               to = as.numeric(c(1, 7, 1, 1, 1, 10, 1,2,log10(6),1, 1))))
FixedEffs_Asia$std.error=as.numeric(FixedEffs_Asia$std.error)
FixedEffs_Asia$estimate=as.numeric(FixedEffs_Asia$estimate)

# Compute 95% CIs
Results_Asia = data.frame(lowerCI = exp(FixedEffs_Asia$Scalar_f*(FixedEffs_Asia$estimate -
                                                                   1.96*FixedEffs_Asia$std.error)),
                          mean = exp(FixedEffs_Asia$Scalar_f*(FixedEffs_Asia$estimate)),
                          upperCI = exp(FixedEffs_Asia$Scalar_f*(FixedEffs_Asia$estimate +
                                                                   1.96*FixedEffs_Asia$std.error)))
rownames(Results_Asia) = FixedEffs_Asia$term
```


### Only African patients


```{r}
SM_Impute_List_Africa = list()
for (i in 1:length(SM_Impute_List)){
  SM_Impute_List_Africa[[i]] = filter(SM_Impute_List[[i]], continent == 'Africa')#s,HCT>9)
}
```

We fit the full model with adjustments as specified in the Methods section:
```{r deathmodel_glm_Africa, include=FALSE, echo=TRUE}
if(RUN_MODELS){
  # the model formula
  DAG_fmla = "outcome ~ HCT + LPAR_pct + coma + convulsions + poedema +
                      log2(BUN) + BD + shock + hypoglycaemia + 
                      drug_AS + (1 | country)"
  # fit the model to each dataset
  modList = glmerModList(DAG_fmla, data = SM_Impute_List_Africa,
                         family=binomial,parallel = T) 
  # save the output
  save(modList, file = 'RData/ModelList_Africa.RData')
} else {
  load('RData/ModelList_Africa.RData')
}

FixedEffs_Africa = modelFixedEff(modList)
print(FixedEffs_Africa)
# Compute the overall parameter estimates:
# The scalar multiples to put the ORs on the correct scales
FixedEffs_Africa$Scalar_f = as.numeric(mapvalues(x = FixedEffs_Africa$term, 
                                                 from = c("(Intercept)",
                                                          "BD",
                                                          "coma",
                                                          "convulsions1",
                                                          "drug_AS",
                                                          "HCT",
                                                          "hypoglycaemia",
                                                          "log2(BUN)",
                                                          "LPAR_pct",
                                                          "shock1",
                                                          "poedema1"),
                                                 to = as.numeric(c(1, 7, 1, 1, 1, 10, 1,2,log10(6),1, 1))))
FixedEffs_Africa$std.error=as.numeric(FixedEffs_Africa$std.error)
FixedEffs_Africa$estimate=as.numeric(FixedEffs_Africa$estimate)

# Compute 95% CIs
Results_Africa = data.frame(lowerCI = exp(FixedEffs_Africa$Scalar_f*(FixedEffs_Africa$estimate -
                                                                       1.96*FixedEffs_Africa$std.error)),
                            mean = exp(FixedEffs_Africa$Scalar_f*(FixedEffs_Africa$estimate)),
                            upperCI = exp(FixedEffs_Africa$Scalar_f*(FixedEffs_Africa$estimate +
                                                                       1.96*FixedEffs_Africa$std.error)))
rownames(Results_Africa) = FixedEffs_Africa$term

```



### Plot in comparison the global estimate


```{r}
Results_Africa = Results_Africa[rownames(Results_Africa) %in% c('BD','coma','convulsions1',
                                                                'drug_AS','HCT',
                                                                'log2(BUN)','poedema1',
                                                                'LPAR_pct','shock1'),]
Results_Africa$Names =mapvalues(rownames(Results_Africa),
                                from = c("drug_AS",
                                         "HCT",
                                         "LPAR_pct",
                                         "convulsions1",
                                         "shock1",
                                         "poedema1",
                                         "log2(BUN)",
                                         "BD", 
                                         "coma"),
                                to = c('Artemisinin drug\nversus\nnon Artemisinin drug',
                                       '-10 % points\nabsolute haematocrit\non admission',
                                       'Six fold increase\n in parasitised\nred blood cells',
                                       'Seizures\non admission',
                                       'Shock\non admission',
                                       'Pulmonary\noedema\non admission',
                                       'Two fold increase\nin blood urea\nnitrogen (mmol/L)',
                                       '+7 mEq/L\nbase deficit',
                                       'Coma\non admission'))
Results_Africa['HCT',c("lowerCI","mean","upperCI")] = 1/Results_Africa['HCT',c("lowerCI","mean","upperCI")]


Results_Asia = Results_Asia[rownames(Results_Asia) %in% c('BD','coma','convulsions1',
                                                          'drug_AS','HCT',
                                                          'log2(BUN)','poedema1',
                                                          'LPAR_pct','shock1'),]
Results_Asia$Names =mapvalues(rownames(Results_Asia),
                              from = c("drug_AS",
                                       "HCT",
                                       "LPAR_pct",
                                       "convulsions1",
                                       "shock1",
                                       "poedema1",
                                       "log2(BUN)",
                                       "BD", 
                                       "coma"),
                              to = c('Artemisinin drug\nversus\nnon Artemisinin drug',
                                     '-10 % points\nabsolute haematocrit\non admission',
                                     'Six fold increase\n in parasitised\nred blood cells',
                                     'Seizures\non admission',
                                     'Shock\non admission',
                                     'Pulmonary\noedema\non admission',
                                     'Two fold increase\nin blood urea\nnitrogen (mmol/L)',
                                     '+7 mEq/L\nbase deficit',
                                     'Coma\non admission'))
Results_Asia['HCT',c("lowerCI","mean","upperCI")] = 1/Results_Asia['HCT',c("lowerCI","mean","upperCI")]
```


```{r Results_Comparison}
library(RColorBrewer)
par(bty='n', las=1, mar = c(4,9,2,2))
epsilon = 0.2 * (1/(nrow(Results)-1))

xlims = c(0.45, 5.5)
plot(NA,NA, xlim= log2(xlims), ylim = c(-2*epsilon,1),xaxt='n',
     xlab='', ylab='', yaxt='n')
abline(v= seq(-1,3,by=1),col = "lightgray", lty = "dotted",lwd = par("lwd"))
axis(1, at = log2(c(0.5,1, 2,4)), labels = c(0.5,1, 2,4))
abline(v=0, lty=2, lwd=3, col='red')
yindex =1
ypos = seq(0,1,length.out = nrow(Results))
cols = brewer.pal(8,'Set1')[2:3]
col1 = cols[1] #adjustcolor('blue',alpha.f = .7)
col2 = cols[2] #adjustcolor('green',alpha.f = .7)
 
for(i in 1:nrow(Results)){
  arrows(log2(Results[i,'lowerCI']),ypos[yindex],
         log2(Results[i,'upperCI']),ypos[yindex],
         length=0.0, angle=90, code=3, 
         col = 'black',lwd=2)
  points(log2(Results[i,'mean']),ypos[yindex],pch=18,cex=1)
  
  ind1 = Results_Africa$Names == Results$Names[i]
  ind2 = Results_Asia$Names == Results$Names[i]
  
  arrows(log2(Results_Africa[ind1,'lowerCI']),ypos[yindex]-epsilon,
         log2(Results_Africa[ind1,'upperCI']),ypos[yindex]-epsilon,
         length=0.0, angle=90, code=3, 
         col = col1,lwd=3)
  points(log2(Results_Africa[ind1,'mean']),ypos[yindex]-epsilon,pch=18,cex=2,col=col1)
  
  arrows(log2(Results_Asia[ind2,'lowerCI']),ypos[yindex]+epsilon,
         log2(Results_Asia[ind2,'upperCI']),ypos[yindex]+epsilon,
         length=0.0, angle=90, code=3, 
         col = col2,lwd=3)
  points(log2(Results_Asia[ind2,'mean']),ypos[yindex]+epsilon,pch=18,cex=2,col=col2)
  
  yindex=yindex+1
  
  
}
abline(h=ypos, lty=3)
axis(side = 2, at = ypos, labels = Results$Names,tick=FALSE)
mtext(side=1, line = 2.5, text = 'Adjusted odds ratio')
mtext(side = 3, line = 1, text = 'Increased survival',adj = 0)
mtext(side = 3, line = 1, text = 'Decreased survival',adj = 1)
legend('bottomright', col = c('black',cols), bty='y',bg = 'white',
       legend = c('All data','African data','Asian data'),lwd=3, inset = 0.02)
```

