---
title: "Charactersing effect of anaemia on mortality in severe malaria"
output:
  html_document:
    keep_md: yes
    fig_caption: yes
---

```{r, echo=F}
knitr::opts_chunk$set(cache = TRUE, cache.comments = FALSE, 
                      echo = TRUE, include = TRUE, 
                      fig.width = 7, fig.height = 7,
                      fig.pos = 'H', 
                      dev = 'png', dpi = 300)

RUN_M_IMPUTATION = F
```

# Background

This looks at the severe malaria legacy dataset from MORU

```{r, echo=FALSE}
load('RData/Data.RData')
m$drug_AS = 0
m$drug_AS[m$drug_class=='artemisinin']=1
```




```{r, echo=FALSE, include=FALSE}
library(lme4)
library(mgcv)
require(boot)
require(doParallel)
require(mitools)
require(missForest)
library(plyr)
library(merTools)
```



Data summaries
```{r}
Africa = c('The Gambia','Mozambique','Ghana','Kenya','Nigeria','Tanzania','Uganda','Rwanda','Congo')
Asia = c('Thailand','Vietnam','Bangladesh','Myanmar','India','Indonesia')
writeLines(paste('Children in Africa:',
                 sum(m$AgeInYear < 15 & m$country %in% Africa)))
writeLines(paste('Adults in Africa:',
                 sum(m$AgeInYear >= 15 & m$country %in% Africa)))

writeLines(paste('Children in Asia:',
                 sum(m$AgeInYear < 15 & m$country %in% Asia)))
writeLines(paste('Adults in Asia:',
                 sum(m$AgeInYear >= 15 & m$country %in% Asia)))

```

### Multiple imputation using missForest

```{r}
apply(m,2, function(x) sum(is.na(x)))
```

```{r}
m$LPAR_pct[is.infinite(m$LPAR_pct)] = 0
m$drug_class = as.factor(m$drug_class)
m$poedema = as.factor(m$poedema)
m$coma = as.factor(m$coma)
m$convulsions = as.factor(m$convulsions)
```

```{r, include=FALSE}
if(RUN_M_IMPUTATION){
  registerDoParallel(cores = 6)
  impute_vars = c('shock','coma','LPAR_pct','BD','BUN','poedema','HCT',
                  'convulsions',"AgeInYear",
                  'rr', 'lactate', 'bicarbonate', 'creatinine')
  SM_Impute_List = list()
  K = 210
  for(i in 1:K){
    Imputed_SM_vars = missForest(xmis = m[,impute_vars], decreasing = T,
                                 maxiter = 10, 
                                 variablewise = T,parallelize = 'forests')
    SM_data = Imputed_SM_vars$ximp
    SM_data = cbind(SM_data, m[,c('outcome','studyID','country','drug_AS')])
    SM_Impute_List[[i]] = SM_data
  }
  
  save(SM_Impute_List, file = 'RData/Multiple_Imputed_Datasets.RData')
} else {
  load(file = 'RData/Multiple_Imputed_Datasets.RData')
}
```


# Unadjusted associations for death








```{r deathmodel_glm}

DAG_fmla = "outcome ~ HCT + LPAR_pct + coma + convulsions + poedema + log10(BUN) + BD + shock + drug_AS + (1 | studyID) + (1 | country)"
modList= glmerModList(DAG_fmla, data = SM_Impute_List, family=binomial) 
Effs = modelFixedEff(modList)
Scalar_f = c(1, 5, 1, 1, 1, 10, 1, 1, 1, 1)
Results = data.frame(lowerCI = exp(Scalar_f*(Effs$estimate - 1.96*Effs$std.error)),
                     mean = exp(Scalar_f*(Effs$estimate)),
                     upperCI = exp(Scalar_f*(Effs$estimate + 1.96*Effs$std.error)))
rownames(Results) = Effs$term
Results
modelRandEffStats(modList)


summary(mod_Adj)
```

Some summary functions we need
```{r}
fmed = function(x, ind){
  return(median(x[ind]))
}
f25 = function(x, ind){
  return(quantile(x[ind], probs = 0.25))
}
f75 = function(x, ind){
  return(quantile(x[ind], probs = 0.75))
}
```

Now let's make counterfactual retrodictions of anaemia on death for the patients in the database.

```{r counterfactualMortality}
myquantiles = c(0.025,0.5,0.975) # this is for a 95% CI
ys = 100*predict(mod_Adj, type='response')
BS = boot(data = ys, statistic = fmed, R = 300)
overall_median_mortality = quantile(BS$t, probs = myquantiles)
```



Compute the counterfactual outcomes:

```{r}
#*********HCT
HCT_CFs = array(dim = c(9, length(x_hcts)))
for(i in 1:length(x_hcts)){
  mydata = Complete_Leg_data
  mydata$HCT=x_hcts[i]
  ys = 100*predict(mod_Adj, newdata = mydata,  type='response')
  BSmed = boot(data = ys, statistic = fmed, R = 300)
  BS25 = boot(data = ys, statistic = f25, R = 300)
  BS75 = boot(data = ys, statistic = f75, R = 300)
  
  HCT_CFs[,i] = c(quantile(BS25$t, probs = myquantiles),
                  quantile(BSmed$t, probs = myquantiles),
                  quantile(BS75$t, probs = myquantiles))
}
#*********BD
BD_CFs = array(dim = c(9, length(x_bd)))
for(i in 1:length(x_bd)){
  mydata = Complete_Leg_data
  mydata$BD=x_bd[i]
  ys = 100*predict(mod_Adj, newdata = mydata, type='response')
  BSmed = boot(data = ys, statistic = fmed, R = 300)
  BS25 = boot(data = ys, statistic = f25, R = 300)
  BS75 = boot(data = ys, statistic = f75, R = 300)
  
  BD_CFs[,i] = c(quantile(BS25$t, probs = myquantiles),
                 quantile(BSmed$t, probs = myquantiles),
                 quantile(BS75$t, probs = myquantiles))
}
#*********BUN
BUN_CFs = array(dim = c(9, length(x_BUN)))
for(i in 1:length(x_BUN)){
  mydata = Complete_Leg_data
  mydata$BUN=x_BUN[i]
  ys = 100*predict(mod_Adj, newdata = mydata,  type='response')
  BSmed = boot(data = ys, statistic = fmed, R = 300)
  BS25 = boot(data = ys, statistic = f25, R = 300)
  BS75 = boot(data = ys, statistic = f75, R = 300)
  
  BUN_CFs[,i] = c(quantile(BS25$t, probs = myquantiles),
                  quantile(BSmed$t, probs = myquantiles),
                  quantile(BS75$t, probs = myquantiles))
}
#*********Coma
coma_CFs = array(dim = c(9, length(x_comas)))
for(i in 1:length(x_comas)){
  mydata = Complete_Leg_data
  mydata$coma=x_comas[i]
  ys = 100*predict(mod_Adj, newdata = mydata,  type='response')
  BSmed = boot(data = ys, statistic = fmed, R = 300)
  BS25 = boot(data = ys, statistic = f25, R = 300)
  BS75 = boot(data = ys, statistic = f75, R = 300)
  
  coma_CFs[,i] = c(quantile(BS25$t, probs = myquantiles),
                   quantile(BSmed$t, probs = myquantiles),
                   quantile(BS75$t, probs = myquantiles))
}
#*********poedema
PO_CFs = array(dim = c(9, length(x_poedema)))
for(i in 1:length(x_poedema)){
  mydata = Complete_Leg_data
  mydata$poedema=x_poedema[i]
  ys = 100*predict(mod_Adj, newdata = mydata, type='response')
  BSmed = boot(data = ys, statistic = fmed, R = 300)
  BS25 = boot(data = ys, statistic = f25, R = 300)
  BS75 = boot(data = ys, statistic = f75, R = 300)
  
  PO_CFs[,i] = c(quantile(BS25$t, probs = myquantiles),
                 quantile(BSmed$t, probs = myquantiles),
                 quantile(BS75$t, probs = myquantiles))
}
#*********Convulsions
Conv_CFs = array(dim = c(9, length(x_conv)))
for(i in 1:length(x_conv)){
  mydata = Complete_Leg_data
  mydata$convulsions=x_conv[i]
  ys = 100*predict(mod_Adj, newdata = mydata,  type='response')
  BSmed = boot(data = ys, statistic = fmed, R = 300)
  BS25 = boot(data = ys, statistic = f25, R = 300)
  BS75 = boot(data = ys, statistic = f75, R = 300)
  
  Conv_CFs[,i] = c(quantile(BS25$t, probs = myquantiles),
                   quantile(BSmed$t, probs = myquantiles),
                   quantile(BS75$t, probs = myquantiles))
}
```

The way to interpret this `counterfactual' plot is as follows: suppose that every individual in the dataset was assigned (as in a intervention) a specific haematocrit $X$, what would the resulting per patient probability of death be. Here we summarise these probabilities by the predicted mean probability of death and 80\% predictive intervals.

```{r CounterfactualPlots_GLM, echo=FALSE}
par(las=1, bty='n', mfrow=c(2,3))
ylabel = 'Probability death'
#************************** HCT
plot(x_hcts,HCT_CFs[5,], xlim=c(4,45), ylab=ylabel, 
     xlab='Haematocrit (%)', ylim=c(0,22), lty=1, lwd=2, type='l')
polygon(c(x_hcts, rev(x_hcts)), c(HCT_CFs[4,], rev(HCT_CFs[6,])),
        border = NA, col = rgb(1, 0, 0,0.5))
lines(x_hcts,HCT_CFs[2,], xlim=c(4,45))
polygon(c(x_hcts, rev(x_hcts)), c(HCT_CFs[1,], rev(HCT_CFs[3,])),
        border = NA, col = rgb(1, 0, 0,0.5))
lines(x_hcts,HCT_CFs[8,], xlim=c(4,45))
polygon(c(x_hcts, rev(x_hcts)), c(HCT_CFs[7,], rev(HCT_CFs[9,])),
        border = NA, col = rgb(1, 0, 0,0.5))
lines(x_hcts, rep(overall_median_mortality[2], length(x_hcts)), lwd=1, col='blue',lty=1)
polygon(c(x_hcts, rev(x_hcts)), 
        c(rep(overall_median_mortality[1],length(x_hcts)),
          rep(overall_median_mortality[3], length(x_hcts))),
        border = NA, col = rgb(0, 0, 1,0.5))

#************************** BD
plot(x_bd,BD_CFs[5,], ylab=ylabel, 
     xlab='Base deficit (mEq/L)', ylim=c(0,80), lty=1, lwd=1, type='l')
lines(x_bd, BD_CFs[2,], lty=1, lwd=1)
lines(x_bd, BD_CFs[8,], lty=1, lwd=1)
polygon(c(x_bd, rev(x_bd)), c(BD_CFs[4,], rev(BD_CFs[6,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_bd, rev(x_bd)), c(BD_CFs[1,], rev(BD_CFs[3,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_bd, rev(x_bd)), c(BD_CFs[7,], rev(BD_CFs[9,])),
        border = NA, col = rgb(1, 0, 0,0.5))
# Overall effect
lines(x_bd, rep(overall_median_mortality[2], length(x_bd)), lwd=1, col='blue',lty=1)
polygon(c(x_bd, rev(x_bd)), 
        c(rep(overall_median_mortality[1],length(x_bd)),
          rep(overall_median_mortality[3], length(x_bd))),
        border = NA, col = rgb(0, 0, 1,0.5))

#************************** BUN
plot(log10(x_BUN),BUN_CFs[5,], ylab=ylabel, xaxt='n',
     xlab='Blood urea nitrogen (mmol/L)', ylim=c(0,45), 
     lty=1, lwd=1, type='l')
axis(1, at = log10(c(2,10,100)), labels = c(2,10,100))
lines(log10(x_BUN), BUN_CFs[2,], lty=1, lwd=1)
lines(log10(x_BUN), BUN_CFs[8,], lty=1, lwd=1)
polygon(log10(c(x_BUN, rev(x_BUN))), c(BUN_CFs[4,], rev(BUN_CFs[6,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(log10(c(x_BUN, rev(x_BUN))), c(BUN_CFs[1,], rev(BUN_CFs[3,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(log10(c(x_BUN, rev(x_BUN))), c(BUN_CFs[7,], rev(BUN_CFs[9,])),
        border = NA, col = rgb(1, 0, 0,0.5))
# Overall effect
lines(log10(x_BUN), rep(overall_median_mortality[2], length(x_BUN)), lwd=1, col='blue',lty=1)
polygon(log10(c(x_BUN, rev(x_BUN))), 
        c(rep(overall_median_mortality[1],length(x_BUN)),
          rep(overall_median_mortality[3], length(x_BUN))),
        border = NA, col = rgb(0, 0, 1,0.5))

#************************** Coma
plot(x_comas,coma_CFs[5,], ylab=ylabel, xaxt='n',
     xlab='', ylim=c(0,30), lty=1, lwd=1, type='l')
axis(1, at = 0:1, labels = c('No coma','Coma'))
lines(x_comas, coma_CFs[2,], lty=1, lwd=1)
lines(x_comas, coma_CFs[8,], lty=1, lwd=1)
polygon(c(x_comas, rev(x_comas)), c(coma_CFs[4,], rev(coma_CFs[6,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_comas, rev(x_comas)), c(coma_CFs[1,], rev(coma_CFs[3,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_comas, rev(x_comas)), c(coma_CFs[7,], rev(coma_CFs[9,])),
        border = NA, col = rgb(1, 0, 0,0.5))
# Overall effect
lines(x_comas, rep(overall_median_mortality[2], length(x_comas)), lwd=1, col='blue',lty=1)
polygon(c(x_comas, rev(x_comas)), 
        c(rep(overall_median_mortality[1],length(x_comas)),
          rep(overall_median_mortality[3], length(x_comas))),
        border = NA, col = rgb(0, 0, 1,0.5))

#************************** Convulsions
plot(x_conv,Conv_CFs[5,], ylab=ylabel, xaxt='n',
     xlab='', ylim=c(0,40), lty=1, lwd=1, type='l')
axis(1, at = x_conv, labels = c('No convulsions','Convulsions'))
lines(x_conv, Conv_CFs[2,], lty=1, lwd=1)
lines(x_conv, Conv_CFs[8,], lty=1, lwd=1)
polygon(c(x_conv, rev(x_conv)), c(Conv_CFs[4,], rev(Conv_CFs[6,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_conv, rev(x_conv)), c(Conv_CFs[1,], rev(Conv_CFs[3,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_conv, rev(x_conv)), c(Conv_CFs[7,], rev(Conv_CFs[9,])),
        border = NA, col = rgb(1, 0, 0,0.5))
# Overall effect
lines(x_conv, rep(overall_median_mortality[2], length(x_conv)), lwd=1, col='blue',lty=1)
polygon(c(x_conv, rev(x_conv)), 
        c(rep(overall_median_mortality[1],length(x_conv)),
          rep(overall_median_mortality[3], length(x_conv))),
        border = NA, col = rgb(0, 0, 1,0.5))

#************************** Pulmonary Oedema
plot(x_poedema,PO_CFs[5,], ylab=ylabel, xaxt='n',
     xlab='', ylim=c(0,40), lty=1, lwd=1, type='l')
axis(1, at = x_poedema, labels = c('No PO','PO'))
lines(x_poedema, PO_CFs[2,], lty=1, lwd=1)
lines(x_poedema, PO_CFs[8,], lty=1, lwd=1)
polygon(c(x_poedema, rev(x_poedema)), c(PO_CFs[4,], rev(PO_CFs[6,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_poedema, rev(x_poedema)), c(PO_CFs[1,], rev(PO_CFs[3,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_poedema, rev(x_poedema)), c(PO_CFs[7,], rev(PO_CFs[9,])),
        border = NA, col = rgb(1, 0, 0,0.5))
# Overall effect
lines(x_poedema, rep(overall_median_mortality[2], length(x_poedema)), lwd=1, col='blue',lty=1)
polygon(c(x_poedema, rev(x_poedema)), 
        c(rep(overall_median_mortality[1],length(x_poedema)),
          rep(overall_median_mortality[3], length(x_poedema))),
        border = NA, col = rgb(0, 0, 1,0.5))
```


## More complex GAM model

The GAM model allows for non-linear relationships between certain variables and the outcome.

Here we fit as non-linear the effect of age and haematocrit on mortality.
We add a random effect term for the studyID and a random effect term for the country.
This should be adjusting for a confounder of quality of care (which potentially impacts both general anaemia and outcome in study).

```{r deathmodel_GAM}

DAG_fmla = "outcome ~ HCT + LPAR_pct + coma + convulsions + poedema + log10(BUN) + BD + shock + drug_AS + (1 | studyID) + (1 | country)"
modList= glmerModList(DAG_fmla, data = SM_Impute_List, family=binomial) 
Effs = modelFixedEff(modList)
Scalar_f = c(1, 5, 1, 1, 1, 10, 1, 1, 1, 1)
Results = data.frame(lowerCI = exp(Scalar_f*(Effs$estimate - 1.96*Effs$std.error)),
                     mean = exp(Scalar_f*(Effs$estimate)),
                     upperCI = exp(Scalar_f*(Effs$estimate + 1.96*Effs$std.error)))
rownames(Results) = Effs$term
Results
modelRandEffStats(modList)


summary(mod_Adj)

mod_Adj_GAM = gam(outcome ~ s(HCT) + LPAR_pct + coma + convulsions +
                    poedema + s(log10(BUN)) + s(BD) + shock + 
                    drug_AS + AgeInYear + 
                    s(studyID, bs='re') + s(country, bs='re'),
                  data=Complete_Leg_data, family=binomial)
summary(mod_Adj_GAM)
```


Now we compute the corresponding counterfactual probabilities of death for the dataset for all values of the haematocrit:

```{r }
#*********HCT
HCT_CFs = array(dim = c(9, length(x_hcts)))
for(i in 1:length(x_hcts)){
  mydata = Complete_Leg_data
  mydata$HCT=x_hcts[i]
  preds = predict(mod_Adj_GAM, newdata = mydata,  type='response', se.fit = T)
  ys = 100*preds$fit
  BSmed = boot(data = ys, statistic = fmed, R = 300)
  BS25 = boot(data = ys, statistic = f25, R = 300)
  BS75 = boot(data = ys, statistic = f75, R = 300)
  
  HCT_CFs[,i] = c(quantile(BS25$t, probs = myquantiles),
                  quantile(BSmed$t, probs = myquantiles),
                  quantile(BS75$t, probs = myquantiles))
}
#*********BD
BD_CFs = array(dim = c(9, length(x_bd)))
for(i in 1:length(x_bd)){
  mydata = Complete_Leg_data
  mydata$BD=x_bd[i]
  ys = 100*predict(mod_Adj_GAM, newdata = mydata, type='response')
  BSmed = boot(data = ys, statistic = fmed, R = 300)
  BS25 = boot(data = ys, statistic = f25, R = 300)
  BS75 = boot(data = ys, statistic = f75, R = 300)
  
  BD_CFs[,i] = c(quantile(BS25$t, probs = myquantiles),
                 quantile(BSmed$t, probs = myquantiles),
                 quantile(BS75$t, probs = myquantiles))
}
#*********BUN
BUN_CFs = array(dim = c(9, length(x_BUN)))
for(i in 1:length(x_BUN)){
  mydata = Complete_Leg_data
  mydata$BUN=x_BUN[i]
  ys = 100*predict(mod_Adj_GAM, newdata = mydata,  type='response')
  BSmed = boot(data = ys, statistic = fmed, R = 300)
  BS25 = boot(data = ys, statistic = f25, R = 300)
  BS75 = boot(data = ys, statistic = f75, R = 300)
  
  BUN_CFs[,i] = c(quantile(BS25$t, probs = myquantiles),
                  quantile(BSmed$t, probs = myquantiles),
                  quantile(BS75$t, probs = myquantiles))
}
#*********Coma
coma_CFs = array(dim = c(9, length(x_comas)))
for(i in 1:length(x_comas)){
  mydata = Complete_Leg_data
  mydata$coma=x_comas[i]
  ys = 100*predict(mod_Adj_GAM, newdata = mydata,  type='response')
  BSmed = boot(data = ys, statistic = fmed, R = 300)
  BS25 = boot(data = ys, statistic = f25, R = 300)
  BS75 = boot(data = ys, statistic = f75, R = 300)
  
  coma_CFs[,i] = c(quantile(BS25$t, probs = myquantiles),
                   quantile(BSmed$t, probs = myquantiles),
                   quantile(BS75$t, probs = myquantiles))
}
#*********poedema
PO_CFs = array(dim = c(9, length(x_poedema)))
for(i in 1:length(x_poedema)){
  mydata = Complete_Leg_data
  mydata$poedema=x_poedema[i]
  ys = 100*predict(mod_Adj_GAM, newdata = mydata, type='response')
  BSmed = boot(data = ys, statistic = fmed, R = 300)
  BS25 = boot(data = ys, statistic = f25, R = 300)
  BS75 = boot(data = ys, statistic = f75, R = 300)
  
  PO_CFs[,i] = c(quantile(BS25$t, probs = myquantiles),
                 quantile(BSmed$t, probs = myquantiles),
                 quantile(BS75$t, probs = myquantiles))
}
#*********Convulsions
Conv_CFs = array(dim = c(9, length(x_conv)))
for(i in 1:length(x_conv)){
  mydata = Complete_Leg_data
  mydata$convulsions=x_conv[i]
  ys = 100*predict(mod_Adj_GAM, newdata = mydata,  type='response')
  BSmed = boot(data = ys, statistic = fmed, R = 300)
  BS25 = boot(data = ys, statistic = f25, R = 300)
  BS75 = boot(data = ys, statistic = f75, R = 300)
  
  Conv_CFs[,i] = c(quantile(BS25$t, probs = myquantiles),
                   quantile(BSmed$t, probs = myquantiles),
                   quantile(BS75$t, probs = myquantiles))
}
```

We see that the effect of haematocrit on mortality is non-linear under this model: below 20 is protective, above 20 plateaus out:

```{r counterfactualPlots_GAM, echo=FALSE}
par(las=1, bty='n', mfrow=c(2,3))
ylabel = 'Probability death'
#************************** HCT
plot(x_hcts,HCT_CFs[5,], xlim=c(4,45), ylab=ylabel, 
     xlab='Haematocrit (%)', ylim=c(0,22), lty=1, lwd=2, type='l')
polygon(c(x_hcts, rev(x_hcts)), c(HCT_CFs[4,], rev(HCT_CFs[6,])),
        border = NA, col = rgb(1, 0, 0,0.5))
lines(x_hcts,HCT_CFs[2,], xlim=c(4,45))
polygon(c(x_hcts, rev(x_hcts)), c(HCT_CFs[1,], rev(HCT_CFs[3,])),
        border = NA, col = rgb(1, 0, 0,0.5))
lines(x_hcts,HCT_CFs[8,], xlim=c(4,45))
polygon(c(x_hcts, rev(x_hcts)), c(HCT_CFs[7,], rev(HCT_CFs[9,])),
        border = NA, col = rgb(1, 0, 0,0.5))
lines(x_hcts, rep(overall_median_mortality[2], length(x_hcts)), lwd=1, col='blue',lty=1)
polygon(c(x_hcts, rev(x_hcts)), 
        c(rep(overall_median_mortality[1],length(x_hcts)),
          rep(overall_median_mortality[3], length(x_hcts))),
        border = NA, col = rgb(0, 0, 1,0.5))

#************************** BD
plot(x_bd,BD_CFs[5,], ylab=ylabel, 
     xlab='Base deficit (mEq/L)', ylim=c(0,80), lty=1, lwd=1, type='l')
lines(x_bd, BD_CFs[2,], lty=1, lwd=1)
lines(x_bd, BD_CFs[8,], lty=1, lwd=1)
polygon(c(x_bd, rev(x_bd)), c(BD_CFs[4,], rev(BD_CFs[6,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_bd, rev(x_bd)), c(BD_CFs[1,], rev(BD_CFs[3,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_bd, rev(x_bd)), c(BD_CFs[7,], rev(BD_CFs[9,])),
        border = NA, col = rgb(1, 0, 0,0.5))
# Overall effect
lines(x_bd, rep(overall_median_mortality[2], length(x_bd)), lwd=1, col='blue',lty=1)
polygon(c(x_bd, rev(x_bd)), 
        c(rep(overall_median_mortality[1],length(x_bd)),
          rep(overall_median_mortality[3], length(x_bd))),
        border = NA, col = rgb(0, 0, 1,0.5))

#************************** BUN
plot(log10(x_BUN),BUN_CFs[5,], ylab=ylabel, xaxt='n',
     xlab='Blood urea nitrogen (mmol/L)', ylim=c(0,45), 
     lty=1, lwd=1, type='l')
axis(1, at = log10(c(2,10,100)), labels = c(2,10,100))
lines(log10(x_BUN), BUN_CFs[2,], lty=1, lwd=1)
lines(log10(x_BUN), BUN_CFs[8,], lty=1, lwd=1)
polygon(log10(c(x_BUN, rev(x_BUN))), c(BUN_CFs[4,], rev(BUN_CFs[6,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(log10(c(x_BUN, rev(x_BUN))), c(BUN_CFs[1,], rev(BUN_CFs[3,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(log10(c(x_BUN, rev(x_BUN))), c(BUN_CFs[7,], rev(BUN_CFs[9,])),
        border = NA, col = rgb(1, 0, 0,0.5))
# Overall effect
lines(log10(x_BUN), 
      rep(overall_median_mortality[2], length(x_BUN)), 
      lwd=1, col='blue',lty=1)
polygon(log10(c(x_BUN, rev(x_BUN))), 
        c(rep(overall_median_mortality[1],length(x_BUN)),
          rep(overall_median_mortality[3], length(x_BUN))),
        border = NA, col = rgb(0, 0, 1,0.5))

#************************** Coma
plot(x_comas,coma_CFs[5,], ylab=ylabel, xaxt='n',
     xlab='', ylim=c(0,30), lty=1, lwd=1, type='l')
axis(1, at = 0:1, labels = c('No coma','Coma'))
lines(x_comas, coma_CFs[2,], lty=1, lwd=1)
lines(x_comas, coma_CFs[8,], lty=1, lwd=1)
polygon(c(x_comas, rev(x_comas)), c(coma_CFs[4,], rev(coma_CFs[6,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_comas, rev(x_comas)), c(coma_CFs[1,], rev(coma_CFs[3,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_comas, rev(x_comas)), c(coma_CFs[7,], rev(coma_CFs[9,])),
        border = NA, col = rgb(1, 0, 0,0.5))
# Overall effect
lines(x_comas, rep(overall_median_mortality[2], length(x_comas)), lwd=1, col='blue',lty=1)
polygon(c(x_comas, rev(x_comas)), 
        c(rep(overall_median_mortality[1],length(x_comas)),
          rep(overall_median_mortality[3], length(x_comas))),
        border = NA, col = rgb(0, 0, 1,0.5))

#************************** Convulsions
plot(x_conv,Conv_CFs[5,], ylab=ylabel, xaxt='n',
     xlab='', ylim=c(0,30), lty=1, lwd=1, type='l')
axis(1, at = x_conv, labels = c('No convulsions','Convulsions'))
lines(x_conv, Conv_CFs[2,], lty=1, lwd=1)
lines(x_conv, Conv_CFs[8,], lty=1, lwd=1)
polygon(c(x_conv, rev(x_conv)), c(Conv_CFs[4,], rev(Conv_CFs[6,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_conv, rev(x_conv)), c(Conv_CFs[1,], rev(Conv_CFs[3,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_conv, rev(x_conv)), c(Conv_CFs[7,], rev(Conv_CFs[9,])),
        border = NA, col = rgb(1, 0, 0,0.5))
# Overall effect
lines(x_conv, rep(overall_median_mortality[2], length(x_conv)), lwd=1, col='blue',lty=1)
polygon(c(x_conv, rev(x_conv)), 
        c(rep(overall_median_mortality[1],length(x_conv)),
          rep(overall_median_mortality[3], length(x_conv))),
        border = NA, col = rgb(0, 0, 1,0.5))

#************************** Pulmonary Oedema
plot(x_poedema,PO_CFs[5,], ylab=ylabel, xaxt='n',
     xlab='', ylim=c(0,30), lty=1, lwd=1, type='l')
axis(1, at = x_poedema, labels = c('No PO','PO'))
lines(x_poedema, PO_CFs[2,], lty=1, lwd=1)
lines(x_poedema, PO_CFs[8,], lty=1, lwd=1)
polygon(c(x_poedema, rev(x_poedema)), c(PO_CFs[4,], rev(PO_CFs[6,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_poedema, rev(x_poedema)), c(PO_CFs[1,], rev(PO_CFs[3,])),
        border = NA, col = rgb(1, 0, 0,0.5))
polygon(c(x_poedema, rev(x_poedema)), c(PO_CFs[7,], rev(PO_CFs[9,])),
        border = NA, col = rgb(1, 0, 0,0.5))
# Overall effect
lines(x_poedema, rep(overall_median_mortality[2], length(x_poedema)), lwd=1, col='blue',lty=1)
polygon(c(x_poedema, rev(x_poedema)), 
        c(rep(overall_median_mortality[1],length(x_poedema)),
          rep(overall_median_mortality[3], length(x_poedema))),
        border = NA, col = rgb(0, 0, 1,0.5))
```


## Model comparison

Which model is better fit in terms of AIC

```{r}
print(AIC(mod_Adj_GAM, mod_Adj))
```

And in terms of deviance

```{r}
print(list(mod_Adj_GAM=deviance(mod_Adj_GAM),
           mod_Adj = deviance(mod_Adj)))
```

