---
title: "Charactersing effect of anaemia on mortality in severe malaria"
output:
  html_document:
    keep_md: yes
    fig_caption: yes
---

```{r, echo=F}
knitr::opts_chunk$set(cache = TRUE, cache.comments = FALSE, 
                      echo = TRUE, include = TRUE, 
                      fig.width = 7, fig.height = 7,
                      fig.pos = 'H', 
                      dev = 'png', dpi = 300)

RUN_MODELS = F
```

# Background

This looks at the severe malaria legacy dataset from MORU

```{r, echo=FALSE}
load('RData/Data.RData')
m$drug_AS = 0
m$drug_AS[m$drug_class=='artemisinin']=1

m$coma[is.na(m$coma) & m$studyID=='AQ'] = 0
```




```{r, echo=FALSE, include=FALSE}
library(lme4)
require(missForest)
library(merTools)
```



Data summaries
```{r}
Africa = c('The Gambia','Mozambique','Ghana','Kenya','Nigeria','Tanzania','Uganda','Rwanda','Congo')
Asia = c('Thailand','Vietnam','Bangladesh','Myanmar','India','Indonesia')
writeLines(paste('Children in Africa:',
                 sum(m$AgeInYear < 15 & m$country %in% Africa)))
writeLines(paste('Adults in Africa:',
                 sum(m$AgeInYear >= 15 & m$country %in% Africa)))

writeLines(paste('Children in Asia:',
                 sum(m$AgeInYear < 15 & m$country %in% Asia)))
writeLines(paste('Adults in Asia:',
                 sum(m$AgeInYear >= 15 & m$country %in% Asia)))

```

### Multiple imputation using linear relationships

The number of missing variables in the pooled data:
```{r}
apply(m,2, function(x) sum(is.na(x)))
```

We make a few data adjustments for the model imputation and fitting:
```{r}
m$LPAR_pct[is.infinite(m$LPAR_pct)] = 0
```

We run the multiple imputation using sequential linear models.

First we setup these linear models. Random effect terms are added for study and country.

```{r, include=FALSE}
# Base deficit from bicarbonate
BD_and_bicarbonate = !is.na(m$BD) & !is.na(m$bicarbonate)
mod_impute1 = lmer(BD ~ bicarbonate + (1 | studyID) + (1 | country), 
                   data= m[BD_and_bicarbonate,])
# Base deficit and lactate
BD_and_lactate = !is.na(m$BD) & !is.na(m$lactate)
mod_impute2 = lmer(BD ~ lactate + (1 | country), 
                   data= m[BD_and_lactate,])
# Base deficit and rr
BD_and_rr = !is.na(m$BD) & !is.na(m$rr)
mod_impute3 = lmer(BD ~ rr + (1 | studyID) + (1 | country), 
                   data= m[BD_and_rr,])

ind1 = is.na(m$BD) & !is.na(m$bicarbonate)
ind2 = !ind1 & is.na(m$BD) & !is.na(m$lactate)
ind3 = !ind1 & !ind2 & is.na(m$BD) & !is.na(m$rr)

# BUN and creatinine
BUN_and_cr = !is.na(m$BUN) & !is.na(m$creatinine)
mod_impute4 = lmer(log(BUN) ~ creatinine + (1 | studyID) + (1 | country), 
                   data= m[BUN_and_cr,])
ind4 = is.na(m$BUN) & !is.na(m$creatinine)

# Age from HCT
HCT_and_Age = !is.na(m$AgeInYear) & !is.na(m$HCT)
mod_impute5 = lmer(HCT ~ AgeInYear + (1 | studyID) + (1 | country),
                   data = m[HCT_and_Age,])
ind5 = is.na(m$HCT) & !is.na(m$AgeInYear)

# marginal distribution of BD to impute remaining
BD_alone = !is.na(m$BD)
mod_impute6 = lmer(BD ~ AgeInYear + (1 | studyID) + (1 | country),
                   data = m[BD_alone,])
ind6 = !ind1 & !ind2 & !ind3 & is.na(m$BD)

# marginal distribution of coma 
Coma_alone = !is.na(m$coma)
mod_impute7 = glmer(coma ~ AgeInYear + hypoglycaemia + (1 | studyID) + (1 | country),
                    family = 'binomial',
                    data = m[Coma_alone,])
ind7 = is.na(m$coma)

# marginal distribution of LPAR_pct 
LPAR_pct_alone = !is.na(m$LPAR_pct)
mod_impute8 = lmer(LPAR_pct ~ AgeInYear + (1 | studyID) + (1 | country),
                   data = m[LPAR_pct_alone,])
ind8 = is.na(m$LPAR_pct)

# marginal distribution of LPAR_pct 
Hypoglycaemia_alone = !is.na(m$hypoglycaemia)
mod_impute9 = glmer(hypoglycaemia ~ 1 + (1 | studyID) + (1 | country),
                    data = m[Hypoglycaemia_alone,], family='binomial')
ind9 = is.na(m$hypoglycaemia)

# marginal distribution of BUN 
BUN_alone = !is.na(m$BUN)
mod_impute10 = lmer(log(BUN) ~ 1 + (1 | studyID) + (1 | country),
                    data = m[BUN_alone,])
ind10 = !ind4 & is.na(m$BUN)
```


```{r}
if(RUN_MODELS){
K_imputations = 500
SM_Impute_List = list()
for (k in 1:K_imputations){
  Imp_data = m
  
  # BD from bicarbonate
  coefs1 = summary(mod_impute1)$coefficients
  Imp_data$BD[ind1] = rnorm(n = sum(ind1), 
                            mean = predict(mod_impute1, newdata = m[ind1,]),
                            sd = coefs1[1,'Std. Error'] + 
                              coefs1[2,'Std. Error']*m$bicarbonate[ind1])
  
  # BD from lactate
  coefs2 = summary(mod_impute2)$coefficients
  Imp_data$BD[ind2] = rnorm(n = sum(ind2), 
                            mean = predict(mod_impute2, newdata = m[ind2,],allow.new.levels=T),
                            sd = coefs2[1,'Std. Error'] + 
                              coefs2[2,'Std. Error']*m$lactate[ind2])
  
  # BD from respiratory rate
  coefs3 = summary(mod_impute3)$coefficients
  Imp_data$BD[ind3] = rnorm(n = sum(ind3), 
                            mean = predict(mod_impute3, newdata = m[ind3,],allow.new.levels=T),
                            sd = coefs3[1,'Std. Error'] + 
                              coefs3[2,'Std. Error']*m$rr[ind3])
  
  # BUN from creatinine
  coefs4 = summary(mod_impute4)$coefficients
  Imp_data$BUN[ind4] = exp(rnorm(n = sum(ind4), 
                                 mean = predict(mod_impute4, newdata = m[ind4,],allow.new.levels=T),
                                 sd = coefs4[1,'Std. Error'] + 
                                   coefs4[2,'Std. Error']*m$creatinine[ind4]))
  
  # HCT from Age
  coefs5 = summary(mod_impute5)$coefficients
  Imp_data$HCT[ind5] = rnorm(n = sum(ind5), 
                             mean = predict(mod_impute5, newdata = m[ind5,],allow.new.levels=T),
                             sd = coefs5[1,'Std. Error'] + 
                               coefs5[2,'Std. Error']*m$AgeInYear[ind5])
  
  # BD from age
  coefs6 = summary(mod_impute6)$coefficients
  Imp_data$BD[ind6] = rnorm(n = sum(ind6), 
                            mean = predict(mod_impute6, newdata = m[ind6,],allow.new.levels=T),
                            sd = coefs6[1,'Std. Error'] + 
                              coefs6[2,'Std. Error']*m$AgeInYear[ind6])
  
  # Coma from hypoglycaemia
  coefs7 = summary(mod_impute7)$coefficients
  Imp_data$coma[ind7] = rbinom(n = sum(ind7), size = 1,
                               predict(mod_impute7, newdata = m[ind7,],
                                       allow.new.levels=T, type='response'))
  
  # Parasitaemia from age
  coefs8 = summary(mod_impute8)$coefficients
  Imp_data$LPAR_pct[ind8] = rnorm(n = sum(ind8), 
                                  mean = predict(mod_impute8, 
                                                 newdata = m[ind8,],
                                                 allow.new.levels=T),
                                  sd = coefs8[1,'Std. Error'] + 
                                    coefs8[2,'Std. Error']*m$AgeInYear[ind8])
  
  # Hypoglycaemia: marginal
  coefs9 = summary(mod_impute9)$coefficients
  Imp_data$hypoglycaemia[ind9] = rbinom(n = sum(ind9),size = 1, 
                                        prob = predict(mod_impute9, 
                                                       newdata = m[ind9,],
                                                       allow.new.levels=T,
                                                       type='response'))
  
  # BUN: marginal
  coefs10 = summary(mod_impute10)$coefficients
  Imp_data$BUN[ind10] = exp(rnorm(n = sum(ind10), 
                                  mean = predict(mod_impute10, newdata = m[ind10,],
                                                 allow.new.levels=T),
                                  sd = coefs10[1,'Std. Error']))
  
  SM_Impute_List[[k]] = Imp_data
  
}
save(SM_Impute_List, file = 'RData/Multiple_Imputed_Datasets.RData')
# Check the number of remaining missing in the last imputed dataset:
vars_explanatory = c('HCT','LPAR_pct','coma' ,'convulsions',
                     'poedema','BUN','BD' ,'shock','hypoglycaemia',
                     'drug_AS','studyID','country')
apply(Imp_data[,vars_explanatory], 2, function(x) sum(is.na(x)))
} else {
  load('RData/Multiple_Imputed_Datasets.RData')
}
```


# Logistic regression model

We fit the full model with adjustments as specified in the Methods section:
```{r deathmodel_glm, include=FALSE, echo=TRUE}
if(RUN_MODELS){
  # the model formula
  DAG_fmla = "outcome ~ HCT + LPAR_pct + coma + convulsions + poedema + 
                        log2(BUN) + BD + shock + hypoglycaemia + 
                        drug_AS + (1 | studyID) + (1 | country)"
  # fit the model to each dataset
  modList= glmerModList(DAG_fmla, data = SM_Impute_List, family=binomial,parallel = T) 
  # save the output
  save(modList, file = 'RData/Models_List.RData')
} else {
  load('RData/Models_List.RData')
}
```

Compute the overall parameter estimates:
```{r}
# extract the fixed and random effects from all the model fits
# These functions then compute the overall estimates
FixedEffs = modelFixedEff(modList)
RandEffs = modelRandEffStats(modList)
print(FixedEffs)
print(RandEffs)
```

Aggregate results for plotting:
```{r}
# The scalar multiples to put the AORs on the correct scales
Scalar_f = c(1, 10, 1, 1, 1, 10, 1, log2(3), 1, 1, 1)
# Compute 95% CIs
Results = data.frame(lowerCI = exp(Scalar_f*(FixedEffs$estimate -
                                               1.96*FixedEffs$std.error)),
                     mean = exp(Scalar_f*(FixedEffs$estimate)),
                     upperCI = exp(Scalar_f*(FixedEffs$estimate +
                                               1.96*FixedEffs$std.error)))
rownames(Results) = FixedEffs$term
```

Make the 'forest' plot:
```{r ForestPlot_SM}
plotting_ind = rownames(Results) %in% c('BD','coma','convulsions1','drug_AS','HCT','log2(BUN)','poedema1')
Results = Results[plotting_ind,]
x_ind = sort.int(Results$mean, index.return = T)$ix
Results = Results[x_ind,]
par(bty='n', las=1, mar = c(3,9,2,2))

Y_Labels = c('+10 mEq/L\nbase deficit',
             'Coma\non admission',
             'Seizures\non admission',
             'Artemisinin drug\nversus\nnon Artemisinin drug',
             '-10 % points\nabsolute haematocrit\non admission',
             '3 fold increase\nin blood urea\nnitrogen (mmol/L)',
             'Pulmonary\nOedema\non admission')
Y_Labels = Y_Labels[x_ind]

xlims = c(0.5, 4.5)
plot(NA,NA, xlim= log2(xlims), ylim = c(0,1),xaxt='n',
     xlab='', ylab='', yaxt='n')
axis(1, at = log2(c(0.5,1, 2,4)), labels = c(0.5,1, 2,4))
abline(v=0, lty=2, lwd=3, col='red')
yindex =1
ypos = seq(0,1,length.out = sum(plotting_ind))

Results['HCT',] = 1/Results['HCT',]
for(i in 1:nrow(Results)){
  arrows(log2(Results[i,'lowerCI']),ypos[yindex],
         log2(Results[i,'upperCI']),ypos[yindex],
         length=0.05, angle=90, code=3, 
         col = 'black',lwd=3)
  points(log2(Results[i,'mean']),ypos[yindex],pch=18,cex=2)
  yindex=yindex+1
  
}
abline(h=ypos, lty=3)
axis(side = 2, at = ypos, labels = Y_Labels,tick=FALSE)
mtext(side=1, line = 2, text = 'Adjusted odds ratio')
mtext(side = 3, line = 1, text = 'Increased survival',adj = 0)
mtext(side = 3, line = 1, text = 'Decreased survival',adj = 1)
```



